---
title: "1DP_05_CellChat"
author: "Domi"
date: "2024-11-27"
output: html_document
editor_options: 
  chunk_output_type: console
---

CellChat analysis 

```{r Install package with defined version, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
remotes::install_version("future", version = "1.19.1") #By default, version 1.34.0 is set, but the function netClustering requires version 1.19.1. Install the required version. 
3
packageVersion("future")
.rs.restartR()
```


#Continue from this step 
```{r, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#Check availability for packges 
# Load reticulate
library(reticulate)

# Check if Python 'umap' module is available
if (!py_module_available("umap")) {
  message("UMAP not available. Setting up Python 3.10 environment...")
  
  # Use Python 3.10
  use_python("C:/Users/Dominika/AppData/Local/Programs/Python/Python310/python.exe", required = TRUE)
  
  # Show Python configuration
  py_config()
  
  # Install required packages in Python 3.10
  py_install("numpy", pip = TRUE)
  py_install("umap-learn", pip = TRUE)
  py_install("numba==0.56.4", pip = TRUE)
  
  # Restart R to apply changes
  message("Installation finished. Restarting R session...")
  .rs.restartR()
  
} else {
  message("UMAP is already available. You can continue.")
}
```


```{r Load packages, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
suppressMessages(library("Matrix"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("CellChat"))
suppressMessages(library("ComplexHeatmap"))
suppressMessages(library("dplyr"))
suppressMessages(library("reshape2"))
suppressMessages(library("ggrepel"))
suppressMessages(library("gridExtra"))
suppressMessages(library("patchwork"))
suppressMessages(library("openxlsx"))
suppressMessages(library("magrittr"))

plots <- list()
col_list <- list()
orders <- list()
source("0F_colours_orders.R", local = F, echo = F, print.eval = F)
source("0F_general_functions.R", local = F, echo = F, print.eval = F)
gene_symbol <- read.table("data/Smartseq3xpress.gene_names.txt", header = TRUE)#File generaetd with zUMIS
```


#For CellCaht analysis DIV cluster was removed because tne number of cells involved in the LiveCells protocol was under the CellChat threshold. For the reason the analysis was done without it. 

```{r CellChat preprocessing, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_mg <- readRDS("ws/1DP_01_seurat_merged_integrated_UMAP_mg_cells.rds")
#Subset for isolation protocols with normalization - RNA assay

seuratmg_woDIV <- subset(seurat_mg, Cell_types_mg_abbr != "DIV")

seuratmg_woDIV@meta.data %<>% mutate("Cell_types_mg_abbr_woDIV" = factor(case_when(
  Cell_types_mg_abbr == "HM1" ~ "HM1",
  Cell_types_mg_abbr == "HM2" ~ "HM2",
  Cell_types_mg_abbr == "INTER" ~ "INTER",
  Cell_types_mg_abbr == "ARM" ~ "ARM",
  Cell_types_mg_abbr == "IRM" ~ "IRM"
  )))
```

#For comparison of protocols, we subset  object for isolation protocols and then individiually processed CellChat analysis 
```{r Divide object for isolation protocols, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#LiveCells
seurat_mg_LC <- subset(seuratmg_woDIV, Project == "Cd11b_small_experiment")
DefaultAssay(seurat_mg_LC) <- "RNA" 
#seurat_mg_LC <- JoinLayers(seurat_mg_LC, assay = "RNA")
all.genes.RNA <- rownames(seurat_mg_LC)
seurat_mg_LC <- NormalizeData(seurat_mg_LC, normalization.method = "LogNormalize", scale.factor = 10000)
seurat_mg_LC <- ScaleData(seurat_mg_LC, features = all.genes.RNA)

#LiveNuclei
seurat_mg_LN <- subset(seuratmg_woDIV, Project == "mouse_live_nuclei")
DefaultAssay(seurat_mg_LN) <- "RNA" 
#seurat_mg_LC <- JoinLayers(seurat_mg_LN, assay = "RNA")
all.genes.RNA <- rownames(seurat_mg_LN)
seurat_mg_LN <- NormalizeData(seurat_mg_LN, normalization.method = "LogNormalize", scale.factor = 10000)
seurat_mg_LN <- ScaleData(seurat_mg_LN, features = all.genes.RNA)

#FixedNuclei
seurat_mg_FN <- subset(seuratmg_woDIV, Project == "mouse_fixed_nuclei")
DefaultAssay(seurat_mg_FN) <- "RNA" 
#seurat_mg_FN <- JoinLayers(seurat_mg_FN, assay = "RNA")
all.genes.RNA <- rownames(seurat_mg_FN)
seurat_mg_FN <- NormalizeData(seurat_mg_FN, normalization.method = "LogNormalize", scale.factor = 10000)
seurat_mg_FN <- ScaleData(seurat_mg_FN, features = all.genes.RNA)



# List of Seurat objects and corresponding sample names
seurat_objects <- list(
  seurat_mg_LC = "LiveCells_MCAO&Sham",
  seurat_mg_LN = "LiveNuclei_MCAO&Sham",
  seurat_mg_FN = "FixedNuclei_MCAO&Sham"
)


# Loop over each Seurat object and sample name
for (seurat_name in names(seurat_objects)) {
  # Assign the current sample name
  sample.name <- seurat_objects[[seurat_name]]
  
  # Load the Seurat object dynamically
  seurat.subset <- get(seurat_name)
  
  #Set idents
  Idents(seurat.subset) <- "Cell_types_mg_abbr_woDIV"
  
  # Create CellChat object and process as per your workflow
  cellchat <- createCellChat(object = seurat.subset, group.by = "Cell_types_mg_abbr_woDIV")
  
   # Drop unused factor levels in cell identities
  cellchat@idents <- droplevels(cellchat@idents)
  
  # Map Ensembl IDs to gene symbols
  cellchat@data@Dimnames[[1]] <- gene_symbol$gene_name[match(cellchat@data@Dimnames[[1]], gene_symbol$gene_id)]
  
  # Set up CellChat database
   cellchat@DB<- CellChatDB.mouse
  
  # Run CellChat analyses
  cellchat <- subsetData(cellchat)
  future::plan("multisession", workers = 2)
  
  cellchat <- identifyOverExpressedGenes(cellchat)
  cellchat <- identifyOverExpressedInteractions(cellchat)
  cellchat <- computeCommunProb(cellchat, population.size = TRUE)
  cellchat <- filterCommunication(cellchat, min.cells = 10)
  cellchat <- computeCommunProbPathway(cellchat)
  cellchat <- aggregateNet(cellchat)
  
  # Save the CellChat object with the sample name in the filename
  saveRDS(cellchat, file = paste0("ws/1DP_04_", sample.name, "_cellchat_processed_RNA.rds"))
  
  print(paste("Processed and saved CellChat for", sample.name))
}
```

#For comparison all individual protocols, the protocols were merged into one object
```{r Prepare merged file for CellChat analysis for all conditions, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#https://rdrr.io/github/sqjin/CellChat/f/tutorial/Comparison_analysis_of_multiple_datasets.Rmd

cellchat_list <- list(
  LiveCells = readRDS("ws/1DP_04_LiveCells_MCAO&Sham_cellchat_processed_RNA.rds"),
  LiveNuclei = readRDS("ws/1DP_04_LiveNuclei_MCAO&Sham_cellchat_processed_RNA.rds"),
  FixedNuclei = readRDS("ws/1DP_04_FixedNuclei_MCAO&Sham_cellchat_processed_RNA.rds")
)

cellchat_merged <- mergeCellChat(cellchat_list, add.names = names(cellchat_list))

cellchat_merged <- computeNetSimilarityPairwise(cellchat_merged, type = "functional")
cellchat_merged <- netEmbedding(cellchat_merged, type = "functional")

# Problem with netClustering about multisession >remotes::install_version("future", version = "1.19.1")

cellchat_merged <- netClustering(cellchat_merged, type = "functional")

saveRDS(cellchat_merged, "ws/1DP_04_CellChat_allconditions.rds")
```

#For comparion only LiveCells and FixedNuclei protocol
```{r Prepare merged file for CellChat analysis for all conditions,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#https://rdrr.io/github/sqjin/CellChat/f/tutorial/Comparison_analysis_of_multiple_datasets.Rmd

cellchat_list <- list(
  LiveCells = readRDS("ws/1DP_04_LiveCells_MCAO&Sham_cellchat_processed_RNA.rds"),
  FixedNuclei = readRDS("ws/1DP_04_FixedNuclei_MCAO&Sham_cellchat_processed_RNA.rds")
)

cellchat_merged <- mergeCellChat(cellchat_list, add.names = names(cellchat_list))

cellchat_merged <- computeNetSimilarityPairwise(cellchat_merged, type = "functional")
cellchat_merged <- netEmbedding(cellchat_merged, type = "functional")

# Problem with netClustering about multisession >remotes::install_version("future", version = "1.19.1")

cellchat_merged <- netClustering(cellchat_merged, type = "functional")

saveRDS(cellchat_merged, "ws/1DP_04_CellChat_LC_FN_conditions.rds")
```

```{r Install package with defined version, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE }
remotes::install_version("future", version = "1.34.0") #By default, version 1.34.0 is set, but the function netClustering requires version 1.19.0. Install the required version. 
3
packageVersion("future")
.rs.restartR()
```