---
title: "1DP_03_GO"
author: "Dominika Dostalova"
date: "2024-11-27"
output: html_document
editor_options: 
  chunk_output_type: console
---

##Gene Ontology (GO) Enrichment Analysis

This analysis identifies biological processes that are significantly enriched among differentially expressed genes (DEGs). By mapping DEGs to GO terms, we gain insights into the functional pathways involved in the conditions under study.
In this case, GO enrichment was performed on genes from the **Scatter Plot**, specifically those with expression values greater than 1 in both **LiveCells** and **FixedNuclei** populations.

**Output files for visualization:**
- `ws/1DP_03_GOanalysis_LCvsFNscatterplot.rds`  
- `results/1DP_03_GOanalysis_LCvsFNscatterplot.xlsx`


```{r Load packages, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
suppressMessages(library("Matrix"))
suppressMessages(library("pheatmap"))
suppressMessages(library("RColorBrewer"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("sctransform"))
suppressMessages(library("DT"))
suppressMessages(library("scales"))
suppressMessages(library("ggrepel"))
suppressMessages(library("dplyr"))
suppressMessages(library("ComplexHeatmap"))
suppressMessages(library("here"))
suppressMessages(library("gridExtra"))
suppressMessages(library("magrittr"))
suppressMessages(library("patchwork"))
suppressMessages(library("purrr"))
suppressMessages(library("org.Mm.eg.db"))
suppressMessages(library("clusterProfiler"))
suppressMessages(library("enrichplot"))
suppressMessages(library("openxlsx"))
plots <- list()
col_list <- list()
orders <- list()
source("0F_general_functions.R", local = F, echo = F, print.eval = F)
source("0F_colours_orders.R", local = F, echo = F, print.eval = F)
gene_symbol <- read.table("data/Smartseq3xpress.gene_names.txt", header = TRUE)
```

#GO for Scatter Plot gene in LiveCells vs FixedNuclei
```{r GO analysis ScatterPlot LiveCells vs FixedNuclei, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE }
topT <- readRDS("ws/1DP_03_top_genes_ARM_file.rds")

gene_list <- list()
gene_list[["up_1_FN"]] <- topT$gene_symbol[topT$FixedNuclei - topT$LiveCells >= 1]
gene_list[["up_1_LC"]] <- topT$gene_symbol[topT$LiveCells - topT$FixedNuclei >= 1]


# List of datasets and a container for the results
datasets <- list(LiveCells = gene_list[["up_1_LC"]], FixedNuclei = gene_list[["up_1_FN"]])
converted_results <- list() # To hold datasets with ENTREZ IDs

# Define thresholds
pval_threshold <- 0.05
logfc_threshold <- 0.1

# Process each dataset
for (name in names(datasets)) {
  dataset <- datasets[[name]]
  
  # Convert cleaned ENSEMBL IDs to ENTREZ IDs
  converted_results[[name]]  <- bitr(dataset, 
                     fromType = "SYMBOL",
                     toType = "ENTREZID",
                     OrgDb = org.Mm.eg.db) 
 converted_results[[name]] <- converted_results[[name]]$ENTREZID
  
}

combined_results <- compareCluster(
  geneCluster = converted_results,
  fun = "enrichGO",
  OrgDb = org.Mm.eg.db,
  keyType = "ENTREZID",
  ont = "BP",
  pvalueCutoff = 0.05
)

pairwise_res <- pairwise_termsim(combined_results)

saveRDS(pairwise_res, "ws/1DP_03_GOanalysis_LCvsFNscatterplot.rds")


#Saving of data in Excel file
# Extract the data.frame from the S4 object
pairwise_res_df <- as.data.frame(pairwise_res@compareClusterResult)

# Filter the data for LiveCells and FixedNuclei
live_cells_data <- subset(pairwise_res_df, Cluster == "LiveCells")
fixed_nuclei_data <- subset(pairwise_res_df, Cluster == "FixedNuclei")

# Create a new workbook
wb <- createWorkbook()

# Add LiveCells data as a sheet
addWorksheet(wb, "LiveCells")
writeData(wb, "LiveCells", live_cells_data)

# Add FixedNuclei data as a sheet
addWorksheet(wb, "FixedNuclei")
writeData(wb, "FixedNuclei", fixed_nuclei_data)

# Save the workbook
saveWorkbook(wb,"results/1DP_03_GOanalysis_LCvsFNscatterplot.xlsx" , overwrite = TRUE)
```