---
title: "scRNA-Seq"
author: "Domi"
date: "09 01 2025"
output: html_document
editor_options: 
  chunk_output_type: console
---

Introduction

This script focuses on processing data specifically for the microglia population. Microglia were extracted from the original dataset based on UMAP clustering, allowing for targeted analysis of their transcriptional profiles.
Subsequent data curration followed similar approach as used for the full dataset (1DP_00_all_cells_raw_data_to_seurat_merged_integrated_UMAP.Rmd), ensuring consistency in processing and analysis.

Here are steps applied after subsetting microglia population from all isolated cells.
**Description**:
        **1. Data Loading and Preprocessing**
            - Importing raw data from multiple samples.
            - subsetting microglia population.
            - Quality control filtering (e.g., feature count thresholds, mitochondrial gene content).
            
        **2. Normalization and Scaling**
            - Log-normalization of raw counts.
            - Data scaling for downstream analysis.
            - SCT normalization
            
        **3. Dimensionality Reduction and Integration**
            - **PCA** (Principal Component Analysis) to reduce dimensionality.
            - **UMAP** (Uniform Manifold Approximation and Projection) for 2D visualization.
			      - **CCAIntegration** - Correction of batch effects to create a unified object.
			
       **4. Clustering and Annotation**
			       - Visualization of clusters in UMAP space.
			       - Identification of cell types based on known marker genes.
			       - Assignment of cell identity labels to clusters.	


Below are the key files prepared for downstream analysis which will used in followed markdown files:

1. **Seurat Object (Microglia Cells)**  
		- **File**: `ws/1DP_01_mg_cells_raw_data_to_seurat_merged_integrated_UMAP.Rmd`  
		- **Description**: Contains micorglia cells captured across isolation protocols (**LiveCells**, **LiveNuclei**, **FixedNuclei**) and integrated using Seurat’s CCAIntegration method.

	2. **Metadata for visualisaiton**  
		- **File**: `ws/1DP_01_mg_seurat_metadata_brm.csv`  
		- **Description**:  Metadata from the Seurat object used for visualisation of non-standard cluster.

	3. **Seurat Object (without cluster 4)**  
		- **File**: `ws/1DP_01_seurat_merged_integrated_UMAP_mg_cells.rds`  
		- **Description**:  Contains microglia cells captured across isolation protocols (**LiveCells**, **LiveNuclei**, **FixedNuclei**) and integrated using Seurat’s CCAIntegration metho but without non-standard cluster.

	4. **Metadata for Microglia Cells without non-standard cluster**  
		- **File**: `ws/1DP_01_mg_seurat_metadata_brm.csv`  
		- **Description**:  Metadata from the Seurat object used for visualisation.

	5. **Cell Population Markers**  
		- **File**: `results/1DP_00_markers_mg_cells.xlsx`  
		- **Description**: Contains marker genes identified for each microglia population across protocols.

Load required package and functions
```{r Load libraries, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
suppressMessages(library("Matrix"))
suppressMessages(library("DropletUtils"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("sctransform"))
suppressMessages(library("dplyr"))
suppressMessages(library("reshape2"))
suppressMessages(library("magrittr"))
suppressMessages(library("openxlsx"))

plots <- list()
col_list <- list()
orders <- list()
source("0F_general_functions.R", local = F, echo = F, print.eval = F)
source("0F_colours_orders.R", local = F, echo = F, print.eval = F)
```

Load the files generated as output from the processing of zUMIs on the raw data
```{r Load files, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#Read output from zUMIs
zumis_output <- readRDS("data/Smartseq3xpress.dgecounts.rds")
#Read all metadata from preprocessed data and subset only to MG clusters
metadata <- read.csv("ws/1DP_00_seurat_all_cells_metadata_for_subset_mg.csv", sep = ",")
rownames(metadata) <- metadata$X
#Remove columns that would be duplicated in the new processing of microglia-only data instead of all-cell data. 
metadata <- metadata[metadata$cca_clusters %in% c("0","1","2","10"), c("Project", "Condition", "Mouse", "Gate", "Cycles", "Age", "Position", "Row", "Column", "Sort_Plate", "Index_Plate",  "FSC.A", "FSC.W", "FSC.H", "SSC.A", "SSC.W", "SSC.H", "FITC.A", "APC.A", "PE.A", "DAPI.A", "Ambiguity", "Exon", "Intergenic", "Intron", "Unmapped", "User", "nreadpairs", "nNontagged", "nUMItag", "UMIfraction", "nGenes", "nGenesUMIs", "nUMIs", "pct_coding", "pct_exon", "nGenes_10000", "nGenesUMIs_10000", "nUMIs_10000", "nGenes_20000", "nGenesUMIs_20000", "nUMIs_20000", "nGenes_50000", "nGenesUMIs_50000", "nUMIs_50000", "nGenes_100000", "nGenesUMIs_100000", "nUMIs_100000", "percent.mt","percent.rib","percent.ERCC", "percent.SIRV", "percent.spike", "cca_clusters")]


#Read table with gene symbols and Ensembl IDs
gene_symbol <- read.table("data/Smartseq3xpress.gene_names.txt", header = TRUE)
```

Prepare seurat object with microglia cells
```{r Preparetion of seurat object, eval=TRUE, message=FALSE, warning=FALSE}
seurat_mg <- CreateSeuratObject(counts = zumis_output$umicount$inex$downsampling$downsampled_20000[,colnames(zumis_output$umicount$inex$downsampling$downsampled_20000) %in% rownames(metadata)])
metadata <- metadata[colnames(seurat_mg),]

seurat_mg@meta.data <- cbind(seurat_mg@meta.data,metadata)

#Remove unnecessary objects
rm(metadata, zumis_output)
```

Subset for individual isolation protocols - dataset is composed of three isolation protocols
```{r Subset for experiments, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_mg_raw <- seurat_mg
seurat_mg <- as.list(NULL)

seurat_mg[["Cd11b_cells"]] <- subset(seurat_mg_raw, subset = Project %in% c("Cd11b_small_experiment"))
seurat_mg[["PU1_fixed_nuclei"]] <- subset(seurat_mg_raw, subset = Project %in% c("mouse_fixed_nuclei"))
seurat_mg[["Cd11b_live_nuclei"]] <- subset(seurat_mg_raw, subset = Project %in% c("mouse_live_nuclei"))
```

SCT Normalization, PCA Calculation, and Data Integration, UMAP
```{r Transfromation, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#Merge objects without integration
seurat_mg_merged <- merge(x = seurat_mg[[1]], y = list(seurat_mg[[2]],seurat_mg[[3]]))

#Filter out spike
seurat_mg_merged <- seurat_mg_merged[!(rownames(seurat_mg_merged) %in% c(grep("^ERCC-",rownames(seurat_mg_merged),value = TRUE),
                                                                grep("^SIRV",rownames(seurat_mg_merged),value = TRUE))), ]

#Normalize using SC Transfomration
seurat_mg_merged <- SCTransform(seurat_mg_merged, verbose = TRUE, return.only.var.genes = FALSE, method = "glmGamPoi",
                                       assay = "RNA",n_genes = NULL, variable.features.n = 3000)

rm(seurat_mg, seurat_mg_raw)

#PCA
DefaultAssay(seurat_mg_merged) <- "SCT" 
seurat_mg_merged <- RunPCA(seurat_mg_merged, npcs = 50, verbose = FALSE, assay = "SCT", reduction.name = "pca")

#Integration - CCA - different types of data
seurat_mg_merged <- IntegrateLayers(seurat_mg_merged, assay = "SCT", method = CCAIntegration, normalization.method = "SCT", orig.reduction = "pca", new.reduction = "integrated.cca", verbose = TRUE)

#Calculate UMAP and clusters for each integration method
seurat_mg_merged <- RunUMAP(seurat_mg_merged, reduction = "integrated.cca",
                         dims = 1:50, verbose = TRUE, reduction.name = "umap.cca")
seurat_mg_merged <- FindNeighbors(seurat_mg_merged, reduction = "integrated.cca",
                               dims = 1:50, verbose = TRUE, graph.name = c("cca.nn","cca.snn"))
seurat_mg_merged <- FindClusters(seurat_mg_merged, resolution = 0.85, verbose = TRUE,
                              cluster.name = "cca_clusters_sbst", graph.name = "cca.snn")
```

Metadata Variable Preparation for Seurat Object
```{r Preparation of columns with needed variables, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# COND_PUBLICATION
seurat_mg_merged@meta.data %<>% mutate("Cond_publication" = factor(
  case_when(
    Condition == "D7" ~ "LiveCells_MCAO",
    Condition == "D7_fixed" ~ "FixedNuclei_MCAO",
    Condition == "D7_live" ~ "LiveNuclei_MCAO",
    Condition == "SHAM" ~ "LiveCells_Sham",
    Condition == "SHAM_fixed" ~ "FixedNuclei_Sham",
    Condition == "SHAM_live" ~ "LiveNuclei_Sham"
  )
))


#check the table 
seurat_mg_merged$Cond_publication %>% table()

# Ensure the 'Cond_publication' column exists and convert it to a factor with the desired order
seurat_mg_merged@meta.data$Cond_publication <- factor(seurat_mg_merged@meta.data$Cond_publication, levels = orders[["Cond_publication"]])
seurat_mg_merged@meta.data$Cond_publication %>% table()


#COND_SHAMvsMCAO
seurat_mg_merged@meta.data %<>% mutate("Cond_SHAMvsMCAO" = factor(
  case_when(
    Cond_publication == "LiveCells_Sham" ~ "Sham",
    Cond_publication == "LiveNuclei_Sham" ~ "Sham",
    Cond_publication == "FixedNuclei_Sham" ~ "Sham",
    Cond_publication == "LiveCells_MCAO" ~ "MCAO",
    Cond_publication == "LiveNuclei_MCAO" ~ "MCAO",
    Cond_publication == "FixedNuclei_MCAO" ~ "MCAO"
  )
))
seurat_mg_merged$Cond_SHAMvsMCAO %>% table()

# Ensure the 'Cond_SHAMvsMCAO' column exists and convert it to a factor with the desired order
seurat_mg_merged@meta.data$Cond_SHAMvsMCAO <- factor(seurat_mg_merged@meta.data$Cond_SHAMvsMCAO , levels = orders[["Cond_SHAMvsMCAO"]])
seurat_mg_merged@meta.data$Cond_SHAMvsMCAO %>%table()



#COND_PREPARATION
seurat_mg_merged@meta.data %<>% mutate("Cond_preparation" = factor(
  case_when(
    Cond_publication == "LiveCells_Sham" ~ "LiveCells",
    Cond_publication == "LiveNuclei_Sham" ~ "LiveNuclei",
    Cond_publication == "FixedNuclei_Sham" ~ "FixedNuclei",
    Cond_publication ==  "LiveCells_MCAO" ~ "LiveCells",
    Cond_publication == "LiveNuclei_MCAO" ~ "LiveNuclei",
    Cond_publication == "FixedNuclei_MCAO" ~ "FixedNuclei"
  )
))
seurat_mg_merged$Cond_preparation %>% table()

# Ensure the 'Cond_preparation' column exists and convert it to a factor with the desired order
seurat_mg_merged@meta.data$Cond_preparation <- factor(seurat_mg_merged@meta.data$Cond_preparation , levels = orders[["Cond_preparation"]])
seurat_mg_merged@meta.data$Cond_preparation %>% table()

seurat_mg_merged@meta.data %<>% mutate("LiveCells" = factor(
  case_when(
    Cond_publication == "LiveCells_Sham" ~ "Sham",
    Cond_publication == "LiveNuclei_Sham" ~ "LiveNuclei + FixedNuclei",
    Cond_publication == "FixedNuclei_Sham" ~ "LiveNuclei + FixedNuclei",
    Cond_publication ==  "LiveCells_MCAO" ~ "MCAO",
    Cond_publication == "LiveNuclei_MCAO" ~ "LiveNuclei + FixedNuclei",
    Cond_publication == "FixedNuclei_MCAO" ~ "LiveNuclei + FixedNuclei"
  )
))

seurat_mg_merged$LiveCells %>% table()

# Ensure the 'LiveCells' column exists and convert it to a factor with the desired order
seurat_mg_merged@meta.data$LiveCells <- factor(seurat_mg_merged@meta.data$LiveCells, levels = orders[["LiveCells"]])
seurat_mg_merged$LiveCells %>% table()

seurat_mg_merged@meta.data %<>% mutate("LiveNuclei" = factor(
  case_when(
    Cond_publication == "LiveNuclei_Sham" ~ "Sham",
    Cond_publication == "LiveNuclei_MCAO" ~ "MCAO",
    Cond_publication == "LiveCells_Sham" ~ "LiveCells + FixedNuclei",
    Cond_publication == "FixedNuclei_Sham" ~ "LiveCells + FixedNuclei",
    Cond_publication ==  "LiveCells_MCAO" ~ "LiveCells + FixedNuclei",
    Cond_publication == "FixedNuclei_MCAO" ~ "LiveCells + FixedNuclei"
  )
))

seurat_mg_merged$LiveNuclei %>% table()

# Ensure the 'LiveNuclei' column exists and convert it to a factor with the desired order
seurat_mg_merged@meta.data$LiveNuclei <- factor(seurat_mg_merged@meta.data$LiveNuclei, levels = orders[["LiveNuclei"]])
seurat_mg_merged$LiveNuclei %>% table()


seurat_mg_merged@meta.data %<>% mutate("FixedNuclei" = factor(
  case_when(
    Cond_publication == "LiveCells_Sham" ~ "LiveCells + LiveNuclei",
    Cond_publication == "LiveNuclei_Sham" ~ "LiveCells + LiveNuclei",
    Cond_publication == "FixedNuclei_Sham" ~ "Sham",
    Cond_publication ==  "LiveCells_MCAO" ~ "LiveCells + LiveNuclei",
    Cond_publication == "LiveNuclei_MCAO" ~ "LiveCells + LiveNuclei",
    Cond_publication == "FixedNuclei_MCAO" ~ "MCAO"
  )
))
seurat_mg_merged@meta.data$FixedNuclei <- factor(seurat_mg_merged@meta.data$FixedNuclei, levels = orders[["FixedNuclei"]])
seurat_mg_merged$FixedNuclei %>% table()

saveRDS(seurat_mg_merged,"ws/1DP_01_seurat_merged_integrated_UMAP_mg_cells_brm.rds")
```

Metadata
```{r Prepare data frame of meta.data, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
metadata_mg <- seurat_mg_merged@meta.data

# View the first few rows of the metadata
head(metadata_mg)

# Ensure metadata is a data frame
metadata_mg_df <- as.data.frame(metadata_mg)

#filtered NA in cca_clusters_sbst
metadata_mg_df_filt <- metadata_mg_df %>%
  filter(!is.na(cca_clusters_sbst))

# Save metadata to a CSV file
write.csv(metadata_mg_df, file = "ws/1DP_01_mg_seurat_metadata_brm.csv", row.names = FALSE)
rm(metadata_mg, metadata_mg_df_filt)
```

Remove cluster with NON-STANDARD properties 
```{r Remove "Debris" from seurat_mg, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_mg_clean <- subset(seurat_mg_merged, subset =  cca_clusters_sbst != "4")
```

SCT Normalization, PCA Calculation, and Data Integration, UMAP
```{r SCTransform Normalization, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#Normalize using SCTransfom
seurat_mg_clean <- SCTransform(seurat_mg_clean, verbose = TRUE, return.only.var.genes = FALSE, method = "glmGamPoi",
                                       assay = "RNA",n_genes = NULL, variable.features.n = 3000, new.assay.name = "SCT_rm")

# Integration
seurat_mg_clean <- IntegrateLayers(seurat_mg_clean, assay = "SCT_rm", method = CCAIntegration, normalization.method = "SCT",
                                 orig.reduction = "pca", new.reduction = "integrated.cca_rm", verbose = TRUE)


# PCA counting}
DefaultAssay(seurat_mg_clean) <- "SCT_rm"
all.genes <- row.names(seurat_mg_clean)
seurat_mg_clean <- RunPCA(seurat_mg_clean, npcs = 50, verbose = FALSE, features = all.genes)

#Checking of PC number
plots[["elb_seurat"]] <- ElbowPlot(seurat_mg_clean, ndims = 50)
plots[["elb_seurat"]]
DimHeatmap(seurat_mg_clean, dims = 1:10, cells = 300, balanced = TRUE) #it doesn't want to be saved in the list of plots

seurat_mg_clean <- RunUMAP(seurat_mg_clean, reduction = "integrated.cca_rm",
                         dims = 1:50, verbose = TRUE, reduction.name = "umap.cca_rm")
seurat_mg_clean <- FindNeighbors(seurat_mg_clean, reduction = "integrated.cca_rm",
                               dims = 1:50, verbose = TRUE, graph.name = c("cca.nn_rm","cca.snn_rm"))
seurat_mg_clean <- FindClusters(seurat_mg_clean, resolution = 0.8, verbose = TRUE, graph.name = "cca.snn_rm", cluster.name = "cca_clusters_sbst_rm")

saveRDS(seurat_mg_clean,"ws/1DP_01_seurat_mg_filtered_rm_SCT.rds")
```


```{r Preparation of columns with needed variables, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# ANNOTATION OF MICROGLIA
seurat_mg_clean@meta.data %<>% mutate("Cell_types_mg_abbr" = factor(case_when(
  cca_clusters_sbst_rm == 0 ~ "HM1",
  cca_clusters_sbst_rm == 1 ~ "HM2",
  cca_clusters_sbst_rm == 2 ~ "INTER",
  cca_clusters_sbst_rm == 3 ~ "ARM",
  cca_clusters_sbst_rm == 4 ~ "IRM",
  cca_clusters_sbst_rm == 5 ~ "DIV"
  )))

#check the table 
seurat_mg_clean$Cell_types_mg_abbr %>% table()

# Ensure the 'Cell_types_mg_abbr' column exists and convert it to a factor with the desired order
seurat_mg_clean$Cell_types_mg_abbr <- factor(seurat_mg_clean@meta.data$Cell_types_mg_abbr, levels = orders[["Cell_types_mg_abbr"]])
seurat_mg_clean$Cell_types_mg_abbr %>% table()

saveRDS(seurat_mg_clean,"ws/1DP_01_seurat_merged_integrated_UMAP_mg_cells.rds")
```

Metadata Variable Preparation for Seurat Object 
Extract metadata for microglia subpopulation after removing of nonstandard cluster
```{r Prepare data frame of meta.data, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}

metadata <- seurat_mg_clean@meta.data

# View the first few rows of the metadata
head(metadata)

# Ensure metadata is a data frame
metadata_df <- as.data.frame(metadata)

#filtered NA in Cell_types_abr
metadata_df_filt <- metadata_df %>%
  filter(!is.na(Cell_types_mg_abbr))

# Save metadata to a CSV file
write.csv(metadata_df, file = "ws/1DP_01_mg_seurat_metadata.csv", row.names = FALSE)
rm(metadata, metadata_df_filt)
```

Microglia markers
```{r Markers, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
DefaultAssay(seurat_mg_clean) <- "RNA" 
Idents(seurat_mg_clean) <- "Cell_types_mg_abbr"

#Normalization of data
seurat_mg_clean <- JoinLayers(seurat_mg_clean)
all.genes.RNA <- rownames(seurat_mg_clean)
seurat_mg_clean <- NormalizeData(seurat_mg_clean, normalization.method = "LogNormalize", scale.factor = 10000)
seurat_mg_clean <- ScaleData(seurat_mg_clean, features = all.genes.RNA)

mg_markers <- FindAllMarkers(seurat_mg_clean, 
                              min.pct = 0.8, only.pos = TRUE, verbose = TRUE, assay = "RNA") %>% split(f = .[["cluster"]]) #split()
#Insert column with gene_symbols
for (subpopulation in names(mg_markers)) {
   sub_data <- mg_markers[[subpopulation]]
  ensembl_ids <- sub_data[, "gene"] 
  gene_symbols <- ENSEMBL_to_Symbol(ensembl_ids)
  sub_data$gene_symbol <- gene_symbols
  mg_markers[[subpopulation]] <- sub_data
}



# Create a workbook
wb <- createWorkbook()

# Loop through each subpopulation in CD11b_markers
for (subpopulation in names(mg_markers)) {
  # Extract the data frame for the subpopulation
  sub_data <- mg_markers[[subpopulation]]
  
  # Add a new sheet named after the subpopulation
  addWorksheet(wb, subpopulation)
  
  # Write the data frame into the sheet
  writeData(wb, subpopulation, sub_data)
}

# Save the workbook to an Excel file
saveWorkbook(wb,"results/1DP_00_markers_mg_cells.xlsx", overwrite = TRUE)
```
