---
title: "1DP_06_library_qc"
author: "Dominika Dostalova"
date: "2026-02-17"
output: html_document
---
Introduction:
The original file was divided into three conditions (LiveCells, LiveNuclei, and FixedNuclei) for comparative analysis. In this section, we describe the preparation process; however, the original file cannot be shared because the file contains data from unrelated experiments that have not yet been published. For this reason, individual CRAM files for each condition (FixedNuclei_raw.cram, LiveNuclei_raw.cram and LiveCells_raw.cram) is provided instead. This analysis is prepared for all obtained cells by the protocols.

# Extract barcodes of cells fro the three condtions - used for subsetting from BAM file in Ubuntu 
# R script
```{r Load data}
suppressMessages(library("Seurat"))
# Object with all cells obtained by isolation methods
seurat <- readRDS("ws/1DP_00_seurat_merged_integrated_UMAP_all_cells.rds")
```

```{r Extract barcodes}
# 1. Get the unique preparation conditions
# Using the column you identified: Cond_preparation
conditions <- unique(seurat@meta.data$Cond_preparation)

# 2. Loop through each condition
for (cond in conditions) {
  
  # Calculate the count (your specific request)
  cell_count <- sum(seurat$Cond_preparation == cond)
  # Extract barcodes for the specific condition
  barcodes <- rownames(seurat@meta.data[seurat@meta.data$Cond_preparation == cond, ])
  
  # Save the file
  file_path <- paste0("ws/", cond, "_barcodes.txt")
  write.table(
    barcodes, 
    file = file_path, 
    quote = FALSE, 
    row.names = FALSE, 
    col.names = FALSE
  )
  
  # 5. NEW: Enhanced message with the count of rows/cells
  message("Condition: ", cond)
  message("--> Total cells detected: ", cell_count)
  message("--> Total cells/barcodes saved: ", length(barcodes))
  message("--> Output file: ", file_path)
  message("---------------------------------------")
}
```

Code for Ubuntu (2204.5.10021.0) terminal
```{bash Prepare the environment}
conda create -n bias_metrics -c bioconda -c conda-forge python=3.9 rseqc samtools trimmomatic seqtk pigz
condata activate bias_metrics
cd /mnt/d/Domi/02_Projects/16_3_bias_metrics/data/01_BAM/
```

```{bash Barcodes in sutable format}
# Convert the extracted format of barcodes into Linux format
sed -i 's/\r//' FixedNuclei_barcodes.txt
sed -i 's/\r//' LiveNuclei_barcodes.txt
sed -i 's/\r//' LiveCells_barcodes.txt
```

```{bash Head of bam file}
# Check the components of BAM file 
# View the first 10 reads to see the variables and tags
samtools view Smartseq3xpress.filtered.Aligned.GeneTagged.UBcorrected.sorted.bam | head -n 10
```

```{bash BAM files for three conditions}
# Extract reads per condtion from the original BAM file - w and w/o TAG sequences 
## 1. Grab only the header
samtools view -H Smartseq3xpress.filtered.Aligned.GeneTagged.UBcorrected.sorted.bam > header.sam
## 2. Grab only the reads matching asked barcodes
samtools view Smartseq3xpress.filtered.Aligned.GeneTagged.UBcorrected.sorted.bam | grep -f FixedNuclei_barcodes.txt > FixedNuclei_body.sam
samtools view Smartseq3xpress.filtered.Aligned.GeneTagged.UBcorrected.sorted.bam | grep -f LiveNuclei_barcodes.txt > LiveNuclei_body.sam
samtools view Smartseq3xpress.filtered.Aligned.GeneTagged.UBcorrected.sorted.bam | grep -f LiveCells_barcodes.txt > LiveCells_body.sam
## 3. Combine them and convert back to BAM
cat header.sam FixedNuclei_body.sam | samtools view -b > FixedNuclei_raw.bam
cat header.sam LiveNuclei_body.sam | samtools view -b > LiveNuclei_raw.bam
cat header.sam LiveCells_body.sam| samtools view -b > LiveCells_raw.bam
```

```{bash BAM files into a compressed from}
samtools view -C --output-fmt-option no_ref=1 FixedNuclei_raw.bam -o FixedNuclei_raw.cram
samtools view -C --output-fmt-option no_ref=1 LiveNuclei_raw.bam -o LiveNuclei_raw.cram
samtools view -C --output-fmt-option no_ref=1 LiveCells_raw.bam -o LiveCells_raw.cram
```

```{bash CRAM file back to BAM file}
samtools view -b FixedNuclei_raw.cram -o FixedNuclei_raw.bam
samtools view -b LiveNuclei_raw.cram -o LiveNuclei_raw.bam
samtools view -b LiveCells_raw.cram -o LiveCells_raw.bam
```

```{bash Statistic about sequences w or w/o TAG}
# Informative statistics about the covarage
## 1. Count total reads
samtools view -c FixedNuclei_raw.bam
## 2. Finds UX:Z: immediately followed by a TAB (indicating no sequence value) - w/o TAG
samtools view FixedNuclei_raw.bam | grep -P "UX:Z:\t" | wc -l
## 3. Finds UB:Z: followed by an actual DNA base - w TAG
samtools view FixedNuclei_raw.bam | grep -P "UB:Z:[ACGT]" | wc -l
```

```{bash BAM files with TAG sequences per condition}
# Extract reads per condtion - w TAG 
## 1. Extract the header first
samtools view -H FixedNuclei_raw.bam > FixedNuclei_header_tag.sam
samtools view -H LiveNuclei_raw.bam > LiveNuclei_header_tag.sam
samtools view -H LiveCells_raw.bam > LiveCells_header_tag.sam
## 2. Filter for reads where UB:Z: is followed by a DNA base [ACGT]
samtools view FixedNuclei_raw.bam | grep -P "UB:Z:[ACGT]" > FixedNuclei_tagged_body.sam
samtools view LiveNuclei_raw.bam | grep -P "UB:Z:[ACGT]" > LiveNuclei_tagged_body.sam
samtools view LiveCells_raw.bam | grep -P "UB:Z:[ACGT]" > LiveCells_tagged_body.sam
## 3. Combine and convert to BAM
cat FixedNuclei_header_tag.sam FixedNuclei_tagged_body.sam | samtools view -b > FixedNuclei_OnlyTagged.bam
cat LiveNuclei_header_tag.sam LiveNuclei_tagged_body.sam | samtools view -b > LiveNuclei_OnlyTagged.bam
cat LiveCells_header_tag.sam LiveCells_tagged_body.sam | samtools view -b > LiveCells_OnlyTagged.bam
```

```{bash BAM files w/o TAG sequences per condition}
# Extract reads per condtion - w/o TAG 
## 1. Extract the header first
samtools view -H FixedNuclei_raw.bam > FixedNuclei_header_nontag.sam
samtools view -H LiveNuclei_raw.bam > LiveNuclei_header_nontag.sam
samtools view -H LiveCells_raw.bam > LiveCells_header_nontag.sam
## 2. Filter for reads where UB:Z: is followed by tab
samtools view FixedNuclei_raw.bam | grep -P "UX:Z:\t" > FixedNuclei_nontagged_body.sam
samtools view LiveNuclei_raw.bam | grep -P "UX:Z:\t" > LiveNuclei_nontagged_body.sam
samtools view LiveCells_raw.bam | grep -P "UX:Z:\t" > LiveCells_nontagged_body.sam
## 3. Combine and convert to BAM
cat FixedNuclei_header_nontag.sam FixedNuclei_nontagged_body.sam | samtools view -b > FixedNuclei_nonTagged.bam
cat LiveNuclei_header_nontag.sam LiveNuclei_nontagged_body.sam | samtools view -b > LiveNuclei_nonTagged.bam
cat LiveCells_header_nontag.sam LiveCells_nontagged_body.sam | samtools view -b > LiveCells_nonTagged.bam
```

```{bash BAM files SPIKE with TAG sequnces per condition}
# SPIKE 
# Search the header (@SQ lines) for common spike-in prefixes
samtools view -H FixedNuclei_raw.bam | grep -E "SIRV|ERCC"
## 1. Extract all spike reads from the TAGGED file (3' Bias metrics)
samtools view -h FixedNuclei_OnlyTagged.bam | grep -E "^@|ERCC-|SIRV" | samtools view -b > FixedNuclei_Spikes_Tagged.bam
samtools view -h LiveNuclei_OnlyTagged.bam | grep -E "^@|ERCC-|SIRV" | samtools view -b > LiveNuclei_Spikes_Tagged.bam
samtools view -h LiveCells_OnlyTagged.bam | grep -E "^@|ERCC-|SIRV" | samtools view -b > LiveCells_Spikes_Tagged.bam
# 2. Extract all spike reads from the nonTAGGED file (Integrity metrics) - internal reads
samtools view -h FixedNuclei_nonTagged.bam | grep -E "^@|ERCC-|SIRV" | samtools view -b > FixedNuclei_Spikes_nonTagged.bam
samtools view -h LiveNuclei_nonTagged.bam | grep -E "^@|ERCC-|SIRV" | samtools view -b > LiveNuclei_Spikes_nonTagged.bam
samtools view -h LiveCells_nonTagged.bam | grep -E "^@|ERCC-|SIRV" | samtools view -b > LiveCells_Spikes_nonTagged.bam
```


Gene Body Coverage (BAM) Analysis:
  - Tool: RSeQC (v5.0.3) via usegalaxy.eu ((https://usegalaxy.eu/?tool_id=toolshed.g2.bx.psu.edu%2Frepos%2Fnilesh%2Frseqc%2Frseqc_geneBody_coverage%2F5.0.3%2Bgalaxy0&version=latest).
  - Reference Models: mm10_RefSeq.bed (https://sourceforge.net/projects/rseqc/files/BED/Mouse_Mus_musculus/mm10_RefSeq.bed.gz/download) - for standard files
                      Spikes_Final_BED12.bed (custom - see script below ) - for files with only spike-in sequences
  - Parameters: Minimum mRNA length = 100.

```{bash Spike Bed file}
# The Galaxy platform requires the BED12 format for this analysis 
awk '$3=="exon" {
    split($10, a, "\""); 
    name=a[2]; 
    start=$4-1; 
    end=$5; 
    len=end-start; 
    print $1"\t"start"\t"end"\t"name"\t0\t"$7"\t"start"\t"end"\t0\t1\t"len"\t0"
}' additional_sequence_annot.gtf > Spikes_Final_BED12.bed
```

The data were initially obtained as individual text files; therefore, we consolidated them into a single, comprehensive file for analysis which is provided.
```{r}
# 1. List all Galaxy text files in your directory
file_paths <- list.files(path = "ws/", 
                         pattern = "Galaxy.*\\.txt", 
                         full.names = TRUE)

# 2. Read the first file to initialize the master data frame
# We read the header to find out how many columns (percentiles) there are
header_check <- read.table(file_paths[1], header = TRUE, sep = "\t", nrows = 1, check.names = FALSE)
master_data <- data.frame(Percentile = 1:(ncol(header_check) - 1))

# 3. Loop through all files and add them as new columns
for (f in file_paths) {
  # Read the file with check.names=FALSE to keep underscores in names
  temp_dat <- read.table(f, header = TRUE, sep = "\t", check.names = FALSE, stringsAsFactors = FALSE)
  
  # Extract the sample name from the first column of the first row
  sample_name <- as.character(temp_dat[1, 1])
  
  # Extract only the numeric coverage values starting from the second column
  coverage_values <- as.numeric(temp_dat[1, -1])
  
  #  Add to the master data frame using the specific name as the header
  master_data[[sample_name]] <- coverage_values
  
}

# 4. Save the combined file
write.table(master_data, "ws/1DP_07_Combined_GeneBody_Coverage.txt", 
            sep = "\t", quote = FALSE, row.names = FALSE)

# See the first few rows to verify the headers
head(master_data)
```
# RNA integrity 
```{bash}
# Filter files with insert size
samtools view FixedNuclei_OnlyTagged.bam | awk '{print ($9 < 0 ? -$9 : $9)}' | grep -v '^0$' > FixedNuclei_isizes_OnlyTagged.txt
samtools view FixedNuclei_nonTagged.bam | awk '{print ($9 < 0 ? -$9 : $9)}' | grep -v '^0$' > FixedNuclei_isizes_nonTagged.txt
samtools view FixedNuclei_Spikes_Tagged.bam | awk '{print ($9 < 0 ? -$9 : $9)}' | grep -v '^0$' > FixedNuclei_isizes_Spikes_Tagged.txt
samtools view FixedNuclei_Spikes_nonTagged.bam | awk '{print ($9 < 0 ? -$9 : $9)}' | grep -v '^0$' > FixedNuclei_isizes_Spikes_nonTagged.txt

samtools view LiveNuclei_OnlyTagged.bam | awk '{print ($9 < 0 ? -$9 : $9)}' | grep -v '^0$' > LiveNuclei_isizes_OnlyTagged.txt
samtools view LiveNuclei_nonTagged.bam | awk '{print ($9 < 0 ? -$9 : $9)}' | grep -v '^0$' > LiveNuclei_isizes_nonTagged.txt
samtools view LiveNuclei_Spikes_Tagged.bam | awk '{print ($9 < 0 ? -$9 : $9)}' | grep -v '^0$' > LiveNuclei_isizes_Spikes_Tagged.txt
samtools view LiveNuclei_Spikes_nonTagged.bam | awk '{print ($9 < 0 ? -$9 : $9)}' | grep -v '^0$' > LiveNuclei_isizes_Spikes_nonTagged.txt

samtools view LiveCells_OnlyTagged.bam | awk '{print ($9 < 0 ? -$9 : $9)}' | grep -v '^0$' > LiveCells_isizes_OnlyTagged.txt
samtools view LiveCells_nonTagged.bam | awk '{print ($9 < 0 ? -$9 : $9)}' | grep -v '^0$' > LiveCells_isizes_nonTagged.txt
samtools view LiveCells_Spikes_Tagged.bam | awk '{print ($9 < 0 ? -$9 : $9)}' | grep -v '^0$' > LiveCells_isizes_Spikes_Tagged.txt
samtools view LiveCells_Spikes_nonTagged.bam | awk '{print ($9 < 0 ? -$9 : $9)}' | grep -v '^0$' > LiveCells_isizes_Spikes_nonTagged.txt
```

