---
title: "SS3x"
author: "Domi"
date: "08 01 2025"
output: html_document
editor_options: 
  chunk_output_type: console
---
This scripts guide you through the analysis of single-cell and single-nucleus RNA sequencing (scRNA-seq and snRNA-seq) data generated from microglial enrichment experiments using two different isolation protocols: LiveCells and FixedNuclei.


This guide will walk you through the preprocessing steps applied after running the **zUMIs** pipeline 
**Description**:
        **1. Data Loading and Preprocessing**
            - Importing raw data from multiple samples.
            - Quality control filtering (e.g., feature count thresholds, mitochondrial gene content).
            
        **2. Normalization and Scaling**
            - Log-normalization of raw counts.
            - Data scaling for downstream analysis.
            - SCT normalization
            
        **3. Dimensionality Reduction and Integration**
            - **PCA** (Principal Component Analysis) to reduce dimensionality.
            - **UMAP** (Uniform Manifold Approximation and Projection) for 2D visualization.
			      - **CCAIntegration** - Correction of batch effects to create a unified object.
			

Below are the key files prepared for downstream analysis which will be used in followed markdown files:

1. **Seurat Object (All Cells)**  
		- **File**: `ws/1DP_00_val_seurat_SCT_normalised_orders.rds`  
		- **Description**: Contains all cells captured across isolation protocols (**LiveCells**, and **FixedNuclei**) and integrated using Seurat’s CCAIntegration method.

---

Load required package and functions
```{r Load libraries, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE }
suppressMessages(library("Matrix")) 
suppressMessages(library("NormExpression"))
suppressMessages(library("DropletUtils"))
suppressMessages(library("pheatmap"))
suppressMessages(library("RColorBrewer"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("sctransform"))
suppressMessages(library("DT"))
suppressMessages(library("scales"))
suppressMessages(library("ggrepel"))
suppressMessages(library("dplyr"))
suppressMessages(library("SeuratWrappers"))
suppressMessages(library("scmap"))
suppressMessages(library("DoubletFinder"))
suppressMessages(library("scCATCH"))
suppressMessages(library("reshape2"))
suppressMessages(library("data.table"))
suppressMessages(library("magrittr"))
suppressMessages(library("openxlsx"))
suppressMessages(library("patchwork"))

plots <- list()
col_list <- list()
orders <- list()
source("0F_general_functions.R", local = F, echo = F, print.eval = F)
source("0F_colours_orders.R", local = F, echo = F, print.eval = F)
```

Load the files generated as output from the processing of zUMIs on the raw data
```{r Load files, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#Read output from zUMIs
zumis_output <- readRDS("data/independent_exp/Smartseq3xpress.dgecounts.rds")
#Read all metadata prepared manually
metadata <- read.csv("data/independent_exp/metadata.csv", sep = ",")
#Read table with gene symbols and IDs
gene_symbol <- read.table("data/independent_exp/Smartseq3xpress.gene_names.txt", header = TRUE)
```

Statistics on coding region and exon coverage across different sequencing depths
```{r Calculation of pct_coding and pct_exon, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
dge <- readRDS("data/independent_exp/Smartseq3xpress.dgecounts.rds")
stats <- fread("data/independent_exp/Smartseq3xpress.readspercell.txt")[!RG %in% "bad"]
stats <- dcast(stats, RG~type, value.var = "N")
stats[, nreadpairs := Ambiguity+Exon+Intergenic+Intron+Unmapped]
reads <- fread("data/independent_exp/Smartseq3xpresskept_barcodes_binned.txt.BCUMIstats.txt")
reads[, UMIfraction := nUMItag/(nNontagged+nUMItag)]
stats <- merge(stats,reads, by.x="RG", by.y="XC")


#Statistic without downsampling
exonic_UMIs <- as.matrix(dge$umicount$inex$all)
exonic_Reads <- as.matrix(dge$readcount$inex$all)

complexity <- data.table( RG = colnames(exonic_UMIs),
                          nGenes = colSums(exonic_Reads>0),
                          nGenesUMIs = colSums(exonic_UMIs>0), 
                          nUMIs = colSums(exonic_UMIs))
stats <- full_join(stats,complexity, by = "RG")
stats[,pct_coding := ((Exon+Intron)/nreadpairs)*100]
stats[,pct_exon := ((Exon)/nreadpairs)*100]


# Statistic downsampled to 1000, 2000, and 5000 reads
exonic_UMIs_1000 <- as.matrix(dge$umicount$inex$downsampling$downsampled_1000)
exonic_Reads_1000 <- as.matrix(dge$readcount$inex$downsampling$downsampled_1000)
complexity_1000 <- data.table( RG = colnames(exonic_UMIs_1000),
                               nGenes_1000 = colSums(exonic_Reads_1000 > 0),
                               nGenesUMIs_1000 = colSums(exonic_UMIs_1000 > 0), 
                               nUMIs_1000 = colSums(exonic_UMIs_1000))

# Repeat for 2000
exonic_UMIs_2000 <- as.matrix(dge$umicount$inex$downsampling$downsampled_2000)
exonic_Reads_2000 <- as.matrix(dge$readcount$inex$downsampling$downsampled_2000)
complexity_2000 <- data.table( RG = colnames(exonic_UMIs_2000),
                               nGenes_2000 = colSums(exonic_Reads_2000 > 0),
                               nGenesUMIs_2000 = colSums(exonic_UMIs_2000 > 0), 
                               nUMIs_2000 = colSums(exonic_UMIs_2000))

# Repeat for 5000
exonic_UMIs_5000 <- as.matrix(dge$umicount$inex$downsampling$downsampled_5000)
exonic_Reads_5000 <- as.matrix(dge$readcount$inex$downsampling$downsampled_5000)
complexity_5000 <- data.table( RG = colnames(exonic_UMIs_5000),
                               nGenes_5000 = colSums(exonic_Reads_5000 > 0),
                               nGenesUMIs_5000 = colSums(exonic_UMIs_5000 > 0), 
                               nUMIs_5000 = colSums(exonic_UMIs_5000))

# Merge the tables
stats <- full_join(stats, complexity_1000, by = "RG")
stats <- full_join(stats, complexity_2000, by = "RG")
stats <- full_join(stats, complexity_5000, by = "RG")


#Statistic downsampled to 10000 reads
exonic_UMIs <- as.matrix(dge$umicount$inex$downsampling$downsampled_10000)
exonic_Reads <- as.matrix(dge$readcount$inex$downsampling$downsampled_10000)
complexity_10000 <- data.table( RG = colnames(exonic_UMIs),
                          nGenes_10000 = colSums(exonic_Reads>0),
                          nGenesUMIs_10000 = colSums(exonic_UMIs>0), 
                          nUMIs_10000 = colSums(exonic_UMIs))
#merge assymetric tables
stats <- full_join(stats,complexity_10000, by = "RG")


#Statistic downsampled to 20000 reads
exonic_UMIs <- as.matrix(dge$umicount$inex$downsampling$downsampled_20000)
exonic_Reads <- as.matrix(dge$readcount$inex$downsampling$downsampled_20000)
complexity_20000 <- data.table( RG = colnames(exonic_UMIs),
                          nGenes_20000 = colSums(exonic_Reads>0),
                          nGenesUMIs_20000 = colSums(exonic_UMIs>0), 
                          nUMIs_20000 = colSums(exonic_UMIs))
#merge assymetric tables
stats <- full_join(stats,complexity_20000, by = "RG")


stats_plot <- stats[stats$RG %in% metadata$Barcode,]
stats_plot <- merge(stats_plot, metadata[, c("Barcode", "Condition")], 
                    by.x = "RG", by.y = "Barcode")
```

Prepare seurat object 
```{r Preparetion of seurat object, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat <- CreateSeuratObject(counts = zumis_output$umicount$inex$downsampling$downsampled_2000)
row.names(metadata) <- metadata$Barcode
metadata <- metadata[row.names(metadata) %in% colnames(seurat),]
metadata <- metadata[colnames(seurat),]

stats <- as.data.frame(stats)
stats <- stats[stats$RG %in% colnames(seurat),]
seurat@meta.data <- cbind(seurat@meta.data,metadata[,c(2:ncol(metadata))])
seurat@meta.data <- cbind(seurat@meta.data,stats[,c(2:ncol(stats))])

#Remove unnecessary objects for next steps
rm(dge, exonic_UMIs, exonic_Reads, complexity, complexity_1000, complexity_2000, complexity_5000, complexity_10000, complexity_20000, metadata, reads, zumis_output, stats)
```

Subset for individual isolation protocols - dataset is composed of two isolation protocols
```{r Subset for individual isolation protocols, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_raw <- seurat
seurat <- as.list(NULL)

seurat[["LiveCells"]] <- subset(seurat_raw, subset = Condition %in% c("LiveCells_MCAO") &  Condition != "empty")
seurat[["FixedNuclei"]] <- subset(seurat_raw, subset = Condition %in% c("FixedNuclei_MCAO") & Condition != "empty")
```

Statistic about percentages of mitochondria, ribosomes, ERCC, SIRV, spike
```{r Calculate percentages of genes, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#Reduce gene_symbol list only to genes detected in experiment - problems with counting of percentages
gene_symbol <- gene_symbol[gene_symbol$gene_id %in% rownames(seurat[[1]]),]


#For each Seurat object in the "seurat" list, this process calculates the percentage of mitochondrial and ribosomal protein genes. The results are stored in two new columns within each Seurat object: "percent.mt" for the percentage of mitochondrial genes and "percent.rib" for the percentage of ribosomal protein genes.
for(i in 1:length(seurat)) {
  seurat[[i]][["percent.mt"]] <- PercentageFeatureSet(seurat[[i]], features = Symbol_to_ENSEMBL(c(grep("^mt-",gene_symbol$gene_name,value = TRUE))))
  seurat[[i]][["percent.rib"]] <- PercentageFeatureSet(seurat[[i]], features = Symbol_to_ENSEMBL(c(grep("^Rp[ls]",gene_symbol$gene_name,value = TRUE))))
  seurat[[i]][["percent.ERCC"]] <- PercentageFeatureSet(seurat[[i]], pattern = "ERCC-")
  seurat[[i]][["percent.SIRV"]] <- PercentageFeatureSet(seurat[[i]], pattern = "SIRV")
  seurat[[i]][["percent.spike"]] <- PercentageFeatureSet(seurat[[i]], pattern = "ERCC-|SIRV")
}
```

Fast check of cutoffs for filtering
```{r Visualisation , eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE} 
# Define the cutoffs for each sample
feature_cutoffs <- c(450, 450)
count_cutoffs <- c(700, 500)
spike_cutoffs <- c(17, 15)
umi_cutoffs <- rep(1, 1)

# Store plots here
plots <- list()

for (i in 1:2) {
  Idents(seurat[[i]]) <- "Condition"
  p1 <- VlnPlot(seurat[[i]], features = "nFeature_RNA", pt.size = 0, split.by = "Condition") +
    NoLegend() + geom_hline(yintercept = feature_cutoffs[i])
  
  p2 <- VlnPlot(seurat[[i]], features = "nCount_RNA", pt.size = 0, split.by = "Condition") +
    scale_y_log10() + NoLegend() + geom_hline(yintercept = count_cutoffs[i])
  
  p3 <- VlnPlot(seurat[[i]], features = "percent.spike", pt.size = 0, assay = "RNA", split.by = "Condition") +
    NoLegend() + geom_hline(yintercept = spike_cutoffs[i])
  
  p4 <- VlnPlot(seurat[[i]], features = "UMIfraction", pt.size = 0, assay = "RNA", split.by = "Condition") +
    NoLegend() + geom_hline(yintercept = umi_cutoffs[i]) + ylim(0.7, 1)
  
  # Combine vertically per sample
 plots[[i]] <- p1 / p2 / p3 / p4
}

# Combine horizontally
combined_plot <- wrap_plots(plots, nrow = 1)
combined_plot

```


SCT Normalization, PCA Calculation, and Data Integration, UMAP
```{r SCTtransform data - normalization, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#Filter unwanted cells
seurat_filter <- as.list(NULL)
seurat_filter[[1]] <- subset(seurat[[1]], subset = percent.spike < 17 & nFeature_RNA > 450 & nCount_RNA > 700) #& UMIfraction > 0.5)
seurat_filter[[2]] <- subset(seurat[[2]], subset = percent.spike < 15 & nFeature_RNA > 450 & nCount_RNA > 500) #& UMIfraction > 0.5)

#Merge objects without integration
seurat_merged <- merge(x = seurat_filter[[1]], y = list(seurat_filter[[2]]))


#Filter out ERCC and SIRV genes
seurat_merged <- seurat_merged[!(rownames(seurat_merged) %in% c(grep("^ERCC-",rownames(seurat[[i]]),value = TRUE),
                                                                grep("^SIRV",rownames(seurat[[i]]),value = TRUE))), ]

#Normalize using SC Transformation
seurat_merged <- SCTransform(seurat_merged, verbose = TRUE, return.only.var.genes = FALSE, method = "glmGamPoi",
                                       assay = "RNA",n_genes = NULL, variable.features.n = 3000, new.assay.name = "SCT")

rm(seurat, i, seurat_filter)

#PCA
DefaultAssay(seurat_merged) <- "SCT"
seurat_merged <- RunPCA(seurat_merged, npcs = 50, verbose = FALSE, assay = "SCT", reduction.name = "pca")

#Integration - CCA - different types of data
seurat_merged <- IntegrateLayers(seurat_merged, assay = "SCT", method = CCAIntegration, normalization.method = "SCT", orig.reduction = "pca", new.reduction = "integrated.cca", verbose = TRUE)

#Calculate UMAP and clusters for each integration method
seurat_merged <- RunUMAP(seurat_merged, reduction = "integrated.cca", dims = 1:50, verbose = TRUE, reduction.name = "umap.cca")
seurat_merged <- FindNeighbors(seurat_merged, reduction = "integrated.cca", dims = 1:50, verbose = TRUE, graph.name = c("cca.nn","cca.snn"))
seurat_merged <- FindClusters(seurat_merged, resolution = 0.3, verbose = TRUE,cluster.name = "cca_clusters", graph.name = "cca.snn")
```


Metadata Variable Preparation for Seurat Object
```{r Preparation of columns with needed variables, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#COND_PUBLICATION
# Change order
seurat_merged@meta.data %<>% mutate("Cond_publication_val" = factor(
  case_when(
    Condition == "LiveCells_MCAO" ~ "LiveCells_MCAO",
    Condition == "FixedNuclei_MCAO" ~ "FixedNuclei_MCAO"
  )
))

seurat_merged@meta.data$Cond_publication_val <- factor(seurat_merged@meta.data$Cond_publication_val , levels = orders[["Cond_publication_val"]])
seurat_merged@meta.data$Cond_publication_val %>% table()

#COND_PREPARATION
seurat_merged@meta.data %<>% mutate("Cond_preparation" = factor(
  case_when(
    Condition ==  "LiveCells_MCAO" ~ "LiveCells",
    Condition == "FixedNuclei_MCAO" ~ "FixedNuclei"
  )
))


#  Change order
seurat_merged@meta.data$Cond_preparation <- factor(seurat_merged@meta.data$Cond_preparation , levels = orders[["Cond_preparation_val"]])
seurat_merged@meta.data$Cond_preparation %>%table()


#Save seurat object
saveRDS(seurat_merged,"ws/1DP_00_val_seurat_SCT_normalised_orders.rds")
```
