---
title: "1DP_00_all_cells_raw_data_into_seurat"
author: "Dominika Dostalova + Pavel Abaffy"
date: "08 01 2025"
output: html_document
editor_options: 
  chunk_output_type: console
---

This scripts guide you through the analysis of single-cell and single-nucleus RNA sequencing (scRNA-seq and snRNA-seq) data generated from microglial enrichment experiments using three different isolation protocols: LiveCells, LiveNuclei, and FixedNuclei.

This guide will walk you through the preprocessing steps applied after running the **zUMIs** pipeline 
**Description**:
        **1. Data Loading and Preprocessing**
            - Importing raw data from multiple samples.
            - Quality control filtering (e.g., feature count thresholds, mitochondrial gene content).
            
        **2. Normalization and Scaling**
            - Log-normalization of raw counts.
            - Data scaling for downstream analysis.
            - SCT normalization
            
        **3. Dimensionality Reduction and Integration**
            - **PCA** (Principal Component Analysis) to reduce dimensionality.
            - **UMAP** (Uniform Manifold Approximation and Projection) for 2D visualization.
			      - **CCAIntegration** - Correction of batch effects to create a unified object.
			
       **4. Clustering and Annotation**
			       - Visualization of clusters in UMAP space.
			       - Identification of cell types based on known marker genes.
			       - Assignment of cell identity labels to clusters.	


Below are the key files prepared for downstream analysis which will used in followed markdown files:

1. **Seurat Object (All Cells)**  
		- **File**: `ws/1DP_00_seurat_merged_integrated_UMAP_all_cells.rds`  
		- **Description**: Contains all cells captured across isolation protocols (**LiveCells**, **LiveNuclei**, **FixedNuclei**) and integrated using Seurat’s CCAIntegration method.

2. **Metadata for Microglia Subsetting**  
		- **File**: `ws/1DP_00_seurat_all_cells_metadata_for_subset_mg.csv`  
		- **Description**: Metadata from the Seurat object used for subsetting microglial populations.

3. **Cell Population Markers**  
		- **File**: `results/1DP_00_markers_all_cells.xlsx`  
		- **Description**: Contains marker genes identified for each cell population across protocols.
---

---


Create folders for saving objects
```{r Folders, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Define folders
folders <- c("ws", "results")

# Create folders if they don't exist
lapply(folders, function(f) if (!dir.exists(f)) dir.create(f))
```


Initialize packages with defined versions 
```{r Packages separation for projects, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
renv::init()#Restore the library state; this will install the exact versions of packages as recorded in the renv.lock file
```

Load required package and functions
```{r Load libraries, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
suppressMessages(library("Matrix")) 
suppressMessages(library("NormExpression"))
suppressMessages(library("DropletUtils"))
suppressMessages(library("pheatmap"))
suppressMessages(library("RColorBrewer"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("sctransform"))
suppressMessages(library("DT"))
suppressMessages(library("scales"))
suppressMessages(library("ggrepel"))
suppressMessages(library("dplyr"))
suppressMessages(library("SeuratWrappers"))
suppressMessages(library("scmap"))
suppressMessages(library("DoubletFinder"))
suppressMessages(library("scCATCH"))
suppressMessages(library("reshape2"))
suppressMessages(library("data.table"))
suppressMessages(library("magrittr"))
suppressMessages(library("openxlsx"))
suppressMessages(library("patchwork"))

plots <- list() #for storing plots
col_list <- list() # for the colors used in the charts
orders <- list() # for variable names in the required order
source("0F_general_functions.R", local = F, echo = F, print.eval = F)
source("0F_colours_orders.R", local = F, echo = F, print.eval = F)
```

Load the files generated as output from the processing of zUMIs on the raw data
```{r Load files, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#Read output from zUMIs
zumis_output <- readRDS("data/Smartseq3xpress.dgecounts.rds")
#Read all metadata prepared manually
metadata <- read.csv("data/metadata.csv", sep = ",")
#Read table with gene symbols and IDs
gene_symbol <- read.table("data/Smartseq3xpress.gene_names.txt", header = TRUE)
```

Statistics on coding region and exon coverage across different sequencing depths
```{r Calculation of pct_coding and pct_exon, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
dge <- readRDS("data/Smartseq3xpress.dgecounts.rds")
stats <- fread("data/Smartseq3xpress.readspercell.txt")[!RG %in% "bad"]
stats <- dcast(stats, RG~type, value.var = "N")
stats[, nreadpairs := Ambiguity+Exon+Intergenic+Intron+Unmapped]
reads <- fread("data/Smartseq3xpresskept_barcodes_binned.txt.BCUMIstats.txt")
reads[, UMIfraction := nUMItag/(nNontagged+nUMItag)]
stats <- merge(stats,reads, by.x="RG", by.y="XC")


#Statistic without downsampling
exonic_UMIs <- as.matrix(dge$umicount$inex$all)
exonic_Reads <- as.matrix(dge$readcount$inex$all)

complexity <- data.table( RG = colnames(exonic_UMIs),
                          nGenes = colSums(exonic_Reads>0),
                          nGenesUMIs = colSums(exonic_UMIs>0), 
                          nUMIs = colSums(exonic_UMIs))
stats <- full_join(stats,complexity, by = "RG")
stats[,pct_coding := ((Exon+Intron)/nreadpairs)*100]
stats[,pct_exon := ((Exon)/nreadpairs)*100]

#Statistic downsampled to 10000 reads
exonic_UMIs <- as.matrix(dge$umicount$inex$downsampling$downsampled_10000)
exonic_Reads <- as.matrix(dge$readcount$inex$downsampling$downsampled_10000)
complexity_10000 <- data.table( RG = colnames(exonic_UMIs),
                          nGenes_10000 = colSums(exonic_Reads>0),
                          nGenesUMIs_10000 = colSums(exonic_UMIs>0), 
                          nUMIs_10000 = colSums(exonic_UMIs))
#merge assymetric tables
stats <- full_join(stats,complexity_10000, by = "RG")


#Statistic downsampled to 20000 reads
exonic_UMIs <- as.matrix(dge$umicount$inex$downsampling$downsampled_20000)
exonic_Reads <- as.matrix(dge$readcount$inex$downsampling$downsampled_20000)
complexity_20000 <- data.table( RG = colnames(exonic_UMIs),
                          nGenes_20000 = colSums(exonic_Reads>0),
                          nGenesUMIs_20000 = colSums(exonic_UMIs>0), 
                          nUMIs_20000 = colSums(exonic_UMIs))
#merge assymetric tables
stats <- full_join(stats,complexity_20000, by = "RG")


#Statistic downsampled to 10000 reads
exonic_UMIs <- as.matrix(dge$umicount$inex$downsampling$downsampled_50000)
exonic_Reads <- as.matrix(dge$readcount$inex$downsampling$downsampled_50000)
complexity_50000 <- data.table( RG = colnames(exonic_UMIs),
                          nGenes_50000 = colSums(exonic_Reads>0),
                          nGenesUMIs_50000 = colSums(exonic_UMIs>0), 
                          nUMIs_50000 = colSums(exonic_UMIs))
#merge assymetric tables
stats <- full_join(stats,complexity_50000, by = "RG")


#Statistic downsampled to 100000 reads
exonic_UMIs <- as.matrix(dge$umicount$inex$downsampling$downsampled_100000)
exonic_Reads <- as.matrix(dge$readcount$inex$downsampling$downsampled_100000)
complexity_100000 <- data.table( RG = colnames(exonic_UMIs),
                          nGenes_100000 = colSums(exonic_Reads>0),
                          nGenesUMIs_100000 = colSums(exonic_UMIs>0), 
                          nUMIs_100000 = colSums(exonic_UMIs))
#merge assymetric tables
stats <- full_join(stats,complexity_100000, by = "RG")


#Statistic downsampled to 200000 reads
exonic_UMIs <- as.matrix(dge$umicount$inex$downsampling$downsampled_200000)
exonic_Reads <- as.matrix(dge$readcount$inex$downsampling$downsampled_200000)
complexity_200000 <- data.table( RG = colnames(exonic_UMIs),
                          nGenes_200000 = colSums(exonic_Reads>0),
                          nGenesUMIs_200000 = colSums(exonic_UMIs>0), 
                          nUMIs_200000 = colSums(exonic_UMIs))
#merge assymetric tables
stats <- full_join(stats,complexity_200000, by = "RG")



#metadata <- metadata[metadata$Project %in% "Cd11b_small_experiment",]

stats_plot <- stats[stats$RG %in% metadata$Barcode,]
```

Prepare seurat object 
```{r Preparetion of seurat object, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat <- CreateSeuratObject(counts = zumis_output$umicount$inex$downsampling$downsampled_20000)
row.names(metadata) <- metadata$Barcode
metadata <- metadata[row.names(metadata) %in% colnames(seurat),]
metadata <- metadata[colnames(seurat),]

stats <- as.data.frame(stats)
stats <- stats[stats$RG %in% colnames(seurat),]
seurat@meta.data <- cbind(seurat@meta.data,metadata[,c(2:ncol(metadata))])
seurat@meta.data <- cbind(seurat@meta.data,stats[,c(2:ncol(stats))])

#Remove unnecessary objects for next steps
rm(dge, exonic_UMIs, exonic_Reads, complexity, complexity_10000, complexity_20000, complexity_50000, complexity_100000, complexity_200000, metadata, reads, zumis_output, stats)
```

Subset for individual isolation protocols - dataset is composed of three isolation protocols
```{r Subset for individual isolation protocols, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_raw <- seurat
seurat <- as.list(NULL)

seurat[["Cd11b_cells"]] <- subset(seurat_raw, subset = Project %in% c("Cd11b_small_experiment") & Condition != "empty")
seurat[["PU1_fixed_nuclei"]] <- subset(seurat_raw, subset = Project %in% c("mouse_fixed_nuclei") & Condition != "empty")
seurat[["Cd11b_live_nuclei"]] <- subset(seurat_raw, subset = Project %in% c("mouse_live_nuclei") & Condition != "empty")
```

Statistic about percentages of mitochondria, ribosomes, ERCC, SIRV, spike
```{r Calculate percentages of genes, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#Reduce gene_symbol list only to genes detected in experiment - problems with counting of percentages
gene_symbol <- gene_symbol[gene_symbol$gene_id %in% rownames(seurat[[1]]),]


#For each Seurat object in the "seurat" list, this process calculates the percentage of mitochondrial and ribosomal protein genes. The results are stored in two new columns within each Seurat object: "percent.mt" for the percentage of mitochondrial genes and "percent.rib" for the percentage of ribosomal protein genes.
for(i in 1:length(seurat)) {
  seurat[[i]][["percent.mt"]] <- PercentageFeatureSet(seurat[[i]], features = Symbol_to_ENSEMBL(c(grep("^mt-",gene_symbol$gene_name,value = TRUE))))
  seurat[[i]][["percent.rib"]] <- PercentageFeatureSet(seurat[[i]], features = Symbol_to_ENSEMBL(c(grep("^Rp[ls]",gene_symbol$gene_name,value = TRUE))))
  seurat[[i]][["percent.ERCC"]] <- PercentageFeatureSet(seurat[[i]], pattern = "ERCC-")
  seurat[[i]][["percent.SIRV"]] <- PercentageFeatureSet(seurat[[i]], pattern = "SIRV")
  seurat[[i]][["percent.spike"]] <- PercentageFeatureSet(seurat[[i]], pattern = "ERCC-|SIRV")
}
```

Fast check of cutoffs for filtering
```{r Violin plots for quality check before filtering, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Define the cutoffs for each sample
feature_cutoffs <- c(1000, 1000, 600)
count_cutoffs <- c(5000, 5000, 2000)
spike_cutoffs <- c(80, 60, 75)
umi_cutoffs <- rep(0.5, 3)

# Store plots here
plots <- list()

for (i in 1:3) {
  p1 <- VlnPlot(seurat[[i]], features = "nFeature_RNA", pt.size = 0, group.by = "Mouse") +
    NoLegend() + geom_hline(yintercept = feature_cutoffs[i])
  
  p2 <- VlnPlot(seurat[[i]], features = "nCount_RNA", pt.size = 0, group.by = "Mouse") +
    scale_y_log10() + NoLegend() + geom_hline(yintercept = count_cutoffs[i])
  
  p3 <- VlnPlot(seurat[[i]], features = "percent.spike", pt.size = 0, group.by = "Mouse", assay = "RNA") +
    NoLegend() + geom_hline(yintercept = spike_cutoffs[i])
  
  p4 <- VlnPlot(seurat[[i]], features = "UMIfraction", pt.size = 0, group.by = "Mouse", assay = "RNA") +
    NoLegend() + geom_hline(yintercept = umi_cutoffs[i]) + ylim(0.7, 1)
  
  # Combine vertically per sample
 plots[[i]] <- p1 / p2 / p3 / p4
}

# Combine horizontally
combined_plot <- wrap_plots(plots, nrow = 1)
combined_plot

```

SCT Normalization, PCA Calculation, and Data Integration, UMAP
```{r SCTtransform data - normalization, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#Filter unwanted cells - thresholds were set in comparison with empty wells
seurat_filter <- as.list(NULL)
seurat_filter[[1]] <- subset(seurat[[1]], subset = percent.spike < 80 & nFeature_RNA > 1000 & nCount_RNA > 5000 & UMIfraction > 0.5)
seurat_filter[[2]] <- subset(seurat[[2]], subset = percent.spike < 60 & nFeature_RNA > 1000 & nCount_RNA > 5000 & UMIfraction > 0.5)
seurat_filter[[3]] <- subset(seurat[[3]], subset = percent.spike < 75 & nFeature_RNA > 600 & nCount_RNA > 2000 & UMIfraction > 0.5)

#Merge objects without integration
seurat_merged <- merge(x = seurat_filter[[1]], y = list(seurat_filter[[2]],seurat_filter[[3]]))


#Filter out ERCC and SIRV genes
seurat_merged <- seurat_merged[!(rownames(seurat_merged) %in% c(grep("^ERCC-",rownames(seurat[[i]]),value = TRUE),
                                                                grep("^SIRV",rownames(seurat[[i]]),value = TRUE))), ]

#Normalize using SC Transformation
seurat_merged <- SCTransform(seurat_merged, verbose = TRUE, return.only.var.genes = FALSE, method = "glmGamPoi",
                                       assay = "RNA",n_genes = NULL, variable.features.n = 3000, new.assay.name = "SCT")

rm(seurat, i, seurat_filter)

#PCA
DefaultAssay(seurat_merged) <- "SCT"
seurat_merged <- RunPCA(seurat_merged, npcs = 50, verbose = FALSE, assay = "SCT", reduction.name = "pca")

#Integration - CCA - different types of data
seurat_merged <- IntegrateLayers(seurat_merged, assay = "SCT", method = CCAIntegration, normalization.method = "SCT", orig.reduction = "pca", new.reduction = "integrated.cca", verbose = TRUE)

#Calculate UMAP and clusters for each integration method
seurat_merged <- RunUMAP(seurat_merged, reduction = "integrated.cca", dims = 1:50, verbose = TRUE, reduction.name = "umap.cca")
seurat_merged <- FindNeighbors(seurat_merged, reduction = "integrated.cca", dims = 1:50, verbose = TRUE, graph.name = c("cca.nn","cca.snn"))
seurat_merged <- FindClusters(seurat_merged, resolution = 0.3, verbose = TRUE,cluster.name = "cca_clusters", graph.name = "cca.snn")

```

Metadata Variable Preparation for Seurat Object
```{r Preparation of columns with needed variables, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Group microglia cluster into one - better for visualisation 
seurat_merged@meta.data %<>% mutate("cca_cl_group" = factor(case_when(
  cca_clusters == 0 ~ "0",
  cca_clusters == 1 ~ "0",
  cca_clusters == 2 ~ "0",
  cca_clusters == 3 ~ "3",
  cca_clusters == 4 ~ "4",
  cca_clusters == 5 ~ "5",
  cca_clusters == 6 ~ "6",
  cca_clusters == 7 ~ "7",
  cca_clusters == 8 ~ "8",
  cca_clusters == 9 ~ "9",
  cca_clusters == 10 ~ "0"
)))

#check the table 
seurat_merged$cca_cl_group %>% table()


# COND_PUBLICATION
seurat_merged@meta.data %<>% mutate("Cond_publication" = factor(
  case_when(
    Condition == "D7" ~ "LiveCells_MCAO",
    Condition == "D7_fixed" ~ "FixedNuclei_MCAO",
    Condition == "D7_live" ~ "LiveNuclei_MCAO",
    Condition == "SHAM" ~ "LiveCells_Sham",
    Condition == "SHAM_fixed" ~ "FixedNuclei_Sham",
    Condition == "SHAM_live" ~ "LiveNuclei_Sham"
  )
))

#check the table 
seurat_merged$Cond_publication %>% table()

# Change order
seurat_merged@meta.data$Cond_publication <- factor(seurat_merged@meta.data$Cond_publication, levels = orders[["Cond_publication"]])
seurat_merged@meta.data$Cond_publication %>% table()


#COND_SHAMvsMCAO
seurat_merged@meta.data %<>% mutate("Cond_SHAMvsMCAO" = factor(
  case_when(
    Cond_publication == "LiveCells_Sham" ~ "Sham",
    Cond_publication == "LiveNuclei_Sham" ~ "Sham",
    Cond_publication == "FixedNuclei_Sham" ~ "Sham",
    Cond_publication == "LiveCells_MCAO" ~ "MCAO",
    Cond_publication == "LiveNuclei_MCAO" ~ "MCAO",
    Cond_publication == "FixedNuclei_MCAO" ~ "MCAO"
  )
))

# Change order
seurat_merged@meta.data$Cond_SHAMvsMCAO <- factor(seurat_merged@meta.data$Cond_SHAMvsMCAO , levels = orders[["Cond_SHAMvsMCAO"]])
seurat_merged@meta.data$Cond_SHAMvsMCAO %>% table()


#COND_PREPARATION
seurat_merged@meta.data %<>% mutate("Cond_preparation" = factor(
  case_when(
    Cond_publication == "LiveCells_Sham" ~ "LiveCells",
    Cond_publication == "LiveNuclei_Sham" ~ "LiveNuclei",
    Cond_publication == "FixedNuclei_Sham" ~ "FixedNuclei",
    Cond_publication ==  "LiveCells_MCAO" ~ "LiveCells",
    Cond_publication == "LiveNuclei_MCAO" ~ "LiveNuclei",
    Cond_publication == "FixedNuclei_MCAO" ~ "FixedNuclei"
  )
))


#  Change order
seurat_merged@meta.data$Cond_preparation <- factor(seurat_merged@meta.data$Cond_preparation , levels = orders[["Cond_preparation"]])
seurat_merged@meta.data$Cond_preparation %>%table()



#CELL_TYPES_ALL_ABBR
seurat_merged@meta.data %<>% mutate("Cell_types_all_abbr" = factor(case_when(
  cca_cl_group == 0 ~ "MG",
  cca_cl_group == 3 ~ "MON",
  cca_cl_group == 4 ~ "OL",
  cca_cl_group == 5 ~ "AST",
  cca_cl_group == 6 ~ "PVM",
  cca_cl_group == 7 ~ "MUR",
  cca_cl_group == 8 ~ "EC",
  cca_cl_group == 9 ~ "NEUT"
)))

#check the table 
seurat_merged$Cell_types_all_abbr %>% table()

#  Change order
seurat_merged@meta.data$Cell_types_all_abbr <- factor(seurat_merged@meta.data$Cell_types_all_abbr, levels = orders[["Cell_types_all_abbr"]])
seurat_merged$Cell_types_all_abbr %>% table()

##LIVECELLS
seurat_merged@meta.data %<>% mutate("LiveCells" = factor(
  case_when(
    Cond_publication == "LiveCells_Sham" ~ "Sham",
    Cond_publication == "LiveNuclei_Sham" ~ "LiveNuclei + FixedNuclei",
    Cond_publication == "FixedNuclei_Sham" ~ "LiveNuclei + FixedNuclei",
    Cond_publication ==  "LiveCells_MCAO" ~ "MCAO",
    Cond_publication == "LiveNuclei_MCAO" ~ "LiveNuclei + FixedNuclei",
    Cond_publication == "FixedNuclei_MCAO" ~ "LiveNuclei + FixedNuclei"
  )
))

seurat_merged$LiveCells %>% table()

#  Change order
seurat_merged@meta.data$LiveCells <- factor(seurat_merged@meta.data$LiveCells, levels = orders[["LiveCells"]])
seurat_merged$LiveCells %>% table()

seurat_merged@meta.data %<>% mutate("LiveNuclei" = factor(
  case_when(
    Cond_publication == "LiveNuclei_Sham" ~ "Sham",
    Cond_publication == "LiveNuclei_MCAO" ~ "MCAO",
    Cond_publication == "LiveCells_Sham" ~ "LiveCells + FixedNuclei",
    Cond_publication == "FixedNuclei_Sham" ~ "LiveCells + FixedNuclei",
    Cond_publication ==  "LiveCells_MCAO" ~ "LiveCells + FixedNuclei",
    Cond_publication == "FixedNuclei_MCAO" ~ "LiveCells + FixedNuclei"
  )
))

seurat_merged$LiveNuclei %>% table()
#  Change order
seurat_merged@meta.data$LiveNuclei <- factor(seurat_merged@meta.data$LiveNuclei, levels = orders[["LiveNuclei"]])
seurat_merged$LiveNuclei %>% table()


seurat_merged@meta.data %<>% mutate("FixedNuclei" = factor(
  case_when(
    Cond_publication == "LiveCells_Sham" ~ "LiveCells + LiveNuclei",
    Cond_publication == "LiveNuclei_Sham" ~ "LiveCells + LiveNuclei",
    Cond_publication == "FixedNuclei_Sham" ~ "Sham",
    Cond_publication ==  "LiveCells_MCAO" ~ "LiveCells + LiveNuclei",
    Cond_publication == "LiveNuclei_MCAO" ~ "LiveCells + LiveNuclei",
    Cond_publication == "FixedNuclei_MCAO" ~ "MCAO"
  )
))
seurat_merged@meta.data$FixedNuclei <- factor(seurat_merged@meta.data$FixedNuclei, levels = orders[["FixedNuclei"]])
seurat_merged$FixedNuclei %>% table()

#Save seurat object
saveRDS(seurat_merged,"ws/1DP_00_seurat_merged_integrated_UMAP_all_cells.rds")
```

Extract metadata
```{r MetaData DataFrame, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# subset metadata
cols <- colnames(seurat_merged@meta.data)[c(2:ncol(seurat_merged@meta.data))]
# save metdata
write.csv(x = seurat_merged@meta.data[cols] , file = "ws/1DP_00_seurat_all_cells_metadata_for_subset_mg.csv", quote = FALSE)
```

Markers of cell populations
```{r Markers, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
DefaultAssay(seurat_merged) <- "RNA" 
Idents(seurat_merged) <- "Cell_types_all_abbr"

#Normalization of data
seurat_merged <- JoinLayers(seurat_merged)
all.genes.RNA <- rownames(seurat_merged)
seurat_merged <- NormalizeData(seurat_merged, normalization.method = "LogNormalize", scale.factor = 10000)
seurat_merged <- ScaleData(seurat_merged, features = all.genes.RNA)

#counting of markers
all_markers <- FindAllMarkers(seurat_merged, 
                              min.pct = 0.8, only.pos = TRUE, verbose = TRUE, assay = "RNA") %>% split(f = .[["cluster"]]) #split()
#Insert column with gene_symbols

for (subpopulation in names(all_markers)) {
   sub_data <- all_markers[[subpopulation]]
  ensembl_ids <- sub_data[, "gene"] 
  gene_symbols <- ENSEMBL_to_Symbol(ensembl_ids)
  sub_data$gene_symbol <- gene_symbols
  all_markers[[subpopulation]] <- sub_data
}



# Create a workbook
wb <- createWorkbook()

# Loop through each subpopulation in CD11b_markers
for (subpopulation in names(all_markers)) {
  # Extract the data frame for the subpopulation
  sub_data <- all_markers[[subpopulation]]
  
  # Add a new sheet named after the subpopulation
  addWorksheet(wb, subpopulation)
  
  # Write the data frame into the sheet
  writeData(wb, subpopulation, sub_data)
}

# Save the workbook to an Excel file
saveWorkbook(wb,"results/1DP_00_markers_all_cells.xlsx", overwrite = TRUE)
```
