---
title: "1DP_04_seurat_microglia_fig3"
author: "Domi"
date: "2024-08-20"
output: html_document
editor_options: 
  chunk_output_type: console
---
Introduction
This script is dedicated to visualizing processed data from **DP_00_all_cells_raw_data_to_seurat_merged_integrated_UMAP.Rmd** and **1DP_05_all_cells_LC_FN_raw_data_to_seurat_merged_integrated_UMAP.Rmd**. The processed data are represented using **violin and box plots** to illustrate reads in coding regions or exons; number of genes or UMIs per cell.



```{r Load packages and lists,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
suppressMessages(library("Matrix"))
suppressMessages(library("RColorBrewer"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("sctransform"))
suppressMessages(library("scales"))
suppressMessages(library("dplyr"))
suppressMessages(library("SeuratWrappers"))
suppressMessages(library("reshape2"))
suppressMessages(library("ComplexHeatmap"))
suppressMessages(library("gridExtra"))
suppressMessages(library("magrittr"))
suppressMessages(library("patchwork"))
suppressMessages(library("tidyr"))
plots <- list()
col_list <- list()
orders <- list()
source("0F_general_functions.R", local = F, echo = F, print.eval = F)
source("0F_colours_orders.R", local = F, echo = F, print.eval = F)
#Read table with gene symbols end IDs
gene_symbol <- read.table("data/Smartseq3xpress.gene_names.txt", header = TRUE)
```

```{r Load seurat object,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat <- readRDS("ws/1DP_00_seurat_merged_integrated_UMAP_all_cells.rds")
```

#Violin plots - w/o Sham and MCAO splitting - illustrate reads in coding regions or exons; number of genes or UMIs per cell
```{r Violin plots for technical parameters_ALL CELLS  w/o SHAM and MCAO,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
DefaultAssay(seurat) <- "RNA"

# Define the features and their corresponding y-axis titles
features_list <- list(
  "pct_coding" = "Reads to coding regions",
  "pct_exon" = "Reads to exons",
  "nFeature_RNA" = "Genes/UMI (20K Reads)",
  "nCount_RNA" = "Total number of UMIs"
)

# List for y-axis titles
y_axis_titles <- list(
  "pct_coding" = "Reads in coding regions (%)",
  "pct_exon" = " Reads in exons (%)",
  "nFeature_RNA" = "Number of genes per cell",
  "nCount_RNA" = " Number of UMIs per cell"
)

# Create the violin plot with box plots and custom y-axis titles
for (feature in names(features_list)) {
  plots[[feature]] <- VlnPlot(seurat, features = feature, 
                              group.by = "Cond_preparation", 
                              pt.size = 0, # Remove individual points
                              cols = col_list[["Cond_preparation"]]) +  
    labs(x = "Conditions", y = y_axis_titles[[feature]]) +  # Manually set y-axis title
    theme_mk +
    remove_grid +
    theme(
      strip.background = element_blank(),
      axis.title.x = element_blank(),
      axis.text.x = element_text(angle = 325, vjust = +0.1, hjust = +0.2)
    ) +
    NoLegend() +
    # Overlay box plot with white fill
    geom_boxplot(width = 0.1, outlier.shape = NA, color = "black", fill = "white")
}

    
layout <- "
ABCD
"
plots[["violin_w/oSM"]] <- plots[["nFeature_RNA"]] + plots[["nCount_RNA"]] + plots[["pct_coding"]] + plots[["pct_exon"]] + plot_layout(design =  layout)

plots[["violin_w/oSM"]]
```

# Fig. 01 E
```{r  Fig01 Violin Plots,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
layout_3rdline <- "
ABCD
"
plots[["3rd_line01"]] <- plots[["nFeature_RNA"]] + plots[["nCount_RNA"]] + plots[["pct_coding"]] + plots[["pct_exon"]] +  plot_layout(design = layout_3rdline, guides = "collect") +
   plot_annotation(
    tag_levels = list(c('E'),  # Use 'a' for automatic labels ('a', 'b', 'c', 'd')
                      theme = theme(
                       plot.tag = element_text(size = 7, face = "plain"))))
          
                     
plots[["3rd_line01"]]

ggsave(
  filename = "results/2V_Fig01_3rdline_violin_wo_ShamMCAO.svg",              
  plot = plots[["3rd_line01"]],          
  device = "svg",                         
  width = 16.9,                             
  height = 4.5,                            
  units = "cm"                          
)
```


#Median values for violin plots w/o Sham and MCAO splitting
```{r Median,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Extract metadata from the Seurat object as a data frame
metadata_df <- seurat@meta.data

# Create a data frame to store medians for all features
medians_df <- data.frame()

# Loop through the features to calculate medians
for (feature in names(features_list)) {
  
  # Check if the feature exists in the metadata
  if (feature %in% colnames(metadata_df)) {
    
    # Calculate the median for each condition
    feature_medians <- metadata_df %>%
      group_by(Cond_preparation) %>%  # Group by relevant conditions
      summarise(median_value = median(.data[[feature]], na.rm = TRUE)) %>%
      mutate(feature = feature)
    
    # Add the result to the medians_df
    medians_df <- bind_rows(medians_df, feature_medians)
    
  } else {
    message(paste("Feature", feature, "not found in metadata"))
  }
}

# View the extracted medians
print(medians_df)
```

#Violin plot w Sham and MCAO splitting
```{r Violin plots for technical parameters,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
metadata_df <- read.csv("ws/1DP_00_seurat_all_cells_metadata_for_subset_mg.csv")
seurat <- readRDS("ws/1DP_00_seurat_merged_integrated_UMAP_all_cells.rds")

DefaultAssay(seurat) <- "RNA"


#correct order of ShamvsMCAO
#Cell_SHAMvsMCAO 
metadata_df$Cond_SHAMvsMCAO <- factor(metadata_df$Cond_SHAMvsMCAO, levels = orders[["Cond_SHAMvsMCAO"]])
metadata_df$Cond_SHAMvsMCAO %>% table() 
# Define the features and their corresponding y-axis titles
features_list <- list(
  "pct_coding" = "Reads to coding regions",
  "pct_exon" = "Reads to exons",
  "nFeature_RNA" = "Genes/UMI (20K Reads)",
  "nCount_RNA" = "Total number of UMIs"
)

# List for y-axis titles
y_axis_titles <- list(
  "pct_coding" = "Reads in coding regions (%)",
  "pct_exon" = " Reads in exons (%)",
  "nFeature_RNA" = "Number of genes per cell",
  "nCount_RNA" = " Number of UMIs per cell"
)


# Create the violin plot with box plots and custom y-axis titles
for (feature in names(features_list)) {
  plots[[paste0(feature, "_w")]] <- VlnPlot(seurat, features = feature, 
                              group.by = "Cond_preparation", 
                              split.by = "Cond_publication",
                              pt.size = 0, # Remove individual points
                              cols = col_list[["Conditions"]]) +  
    labs(x = "Condition",  y = y_axis_titles[[feature]]) +  # Manually set y-axis title
    theme_mk +
    remove_grid +
   facet_grid(. ~ metadata_df$Cond_SHAMvsMCAO, drop = TRUE) + 
    theme(
      strip.background = element_blank(),
       axis.title.x = element_blank(),
      axis.text.x = element_text(angle = 325, vjust = +0.1, hjust = +0.2),
      axis.title.y = element_text(margin = margin(t = -10)),
      plot.title = element_blank()
    ) +
    NoLegend() +
        # Overlay box plot with white fill
    geom_boxplot(width = 0.1, outlier.shape = NA, color = "black", fill = "white")
}

      

layout <- "
AB
CD
"
p <- plots[["pct_coding_w"]] + plots[["pct_exon_w"]] + plots[["nFeature_RNA_w"]] + plots[["nCount_RNA_w"]] +  plot_layout(design =  layout)
p

```
 
#Median values for violin plots w Sham and MCAO splitting
```{r Median,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Extract metadata from the Seurat object as a data frame
metadata_df <- seurat@meta.data

# Create a data frame to store medians for all features
medians_df <- data.frame()

# Loop through the features to calculate medians
for (feature in names(features_list)) {
  
  # Check if the feature exists in the metadata
  if (feature %in% colnames(metadata_df)) {
    
    # Calculate the median for each condition
    feature_medians <- metadata_df %>%
      group_by(Cond_preparation, Cond_publication) %>%  # Group by relevant conditions
      summarise(median_value = median(.data[[feature]], na.rm = TRUE)) %>%
      mutate(feature = feature)
    
    # Add the result to the medians_df
    medians_df <- bind_rows(medians_df, feature_medians)
    
  } else {
    message(paste("Feature", feature, "not found in metadata"))
  }
}

# View the extracted medians
print(medians_df)
```

#SFig 01 A
```{r FigS 03,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
layout_1st_line <- "
AB
CD
"
plots[["violin_wSM"]] <- plots[["nFeature_RNA_w"]] + plots[["nCount_RNA_w"]] + plots[["pct_coding_w"]] + plots[["pct_exon_w"]] +  plot_layout(design = layout_1st_line)  &
  plot_annotation(
    tag_levels = list(c('A', " "), 
                      theme = theme(
                        plot.tag = element_text(size = 7, face = "plain")  
                      )))

plots[["violin_wSM"]]

ggsave(
  filename = "results/2V_FigS03_1stline_ViolinPlots_w_ShamMCAO.svg",             
  plot = plots[["violin_wSM"]],          
  device = "svg",                          
  width = 16.9,                            
  height = 9,                              
  units = "cm"                             
)

```

#Box plot -  genes per cell and UMIs per cell at different downsampling levels
```{r Box plot,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
metadata <- read.csv("ws/1DP_00_seurat_all_cells_metadata_for_subset_mg.csv")
seurat <- readRDS("ws/1DP_00_seurat_merged_integrated_UMAP_all_cells.rds")
DefaultAssay(seurat) <- "RNA"



metadata$Cond_preparation <- factor(metadata$Cond_preparation, levels = orders[["Cond_preparation"]])
metadata$Cond_preparation %>% table()

# Function to calculate the standard error of the mean
sem <- function(x) {
  return(sd(x, na.rm = TRUE) / sqrt(length(na.omit(x))))
}

# Extract and reshape the nGenes data
nGenes_long <- metadata %>%
  select(Cond_preparation, starts_with("nGenes_")) %>%
  pivot_longer(cols = starts_with("nGenes_"), names_to = "read_pair", values_to = "nGenes") %>%
  mutate(read_pair = as.numeric(gsub("nGenes_", "", read_pair))) %>%
  drop_na(nGenes)

# Summarize the nGenes data
nGenes_summary <- nGenes_long %>%
  group_by(Cond_preparation, read_pair) %>%
  summarise(mean_nGenes = mean(nGenes, na.rm = TRUE),
            lower_ci = mean_nGenes - qt(1 - 0.05 / 2, length(nGenes) - 1) * sem(nGenes),
            upper_ci = mean_nGenes + qt(1 - 0.05 / 2, length(nGenes) - 1) * sem(nGenes),
            .groups = 'drop')

# Extract and reshape the nUMIs data
nUMIs_long <- metadata %>%
  select(Cond_preparation, starts_with("nUMIs_")) %>%
  pivot_longer(cols = starts_with("nUMIs_"), names_to = "read_pair", values_to = "nUMIs") %>%
  mutate(read_pair = as.numeric(gsub("nUMIs_", "", read_pair))) %>%
  drop_na(nUMIs)

# Summarize the nUMIs data
nUMIs_summary <- nUMIs_long %>%
  group_by(Cond_preparation, read_pair) %>%
  summarise(mean_nUMIs = mean(nUMIs, na.rm = TRUE),
            lower_ci = mean_nUMIs - qt(1 - 0.05 / 2, length(nUMIs) - 1) * sem(nUMIs),
            upper_ci = mean_nUMIs + qt(1 - 0.05 / 2, length(nUMIs) - 1) * sem(nUMIs),
            .groups = 'drop')

# Custom color palette
custom_colors <- brewer.pal(n = length(unique(nGenes_summary$Cond_preparation)), name = "Set1")


# Custom function to add scientific notation-like labels
add_sci_superscript <- function(x) {
  sapply(x, function(value) {
    numeric_value <- as.numeric(value)
    if (!is.na(numeric_value)) {
      exponent <- floor(log10(numeric_value))
      base <- numeric_value / (10^exponent)
      bquote(.(base) %*% 10^.(exponent))  # Format as base × 10^exponent
    } else {
      value  # Return non-numeric values as-is
    }
  })
}




# Boxplot for nGenes
nGenes_long_filtered <- nGenes_long %>%
  filter(read_pair <= 100000)


plots[["nGenes_long"]] <- ggplot(nGenes_long_filtered, aes(x = as.factor(read_pair), y = nGenes, fill = Cond_preparation)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = col_list[["Cond_preparation"]]) +
  labs(x = "Read pairs", y = "Number of genes per cell") +
  scale_x_discrete(labels = add_sci_superscript)+
  coord_cartesian(ylim = c(0,7650))+
  theme_mk +
  remove_grid+
 NoLegend()

plots[["nGenes_long"]]


# Boxplot for nUMIs

nUMIs_long_filtered <- nUMIs_long %>%
  filter(read_pair <= 100000)

plots[["nUMIs_long"]] <- ggplot(nUMIs_long_filtered, aes(x = as.factor(read_pair), y = nUMIs, fill = Cond_preparation)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  scale_fill_manual(values = col_list[["Cond_preparation"]]) +
  labs(x = "Read pairs", y = "Number of UMIs per cell")+
  scale_x_discrete(labels = add_sci_superscript)+
  scale_y_continuous(limits = c(0,44000))+
  theme_mk+
  remove_grid +
  NoLegend()
  

plots[["nUMIs_long"]]
```

#SFig 01 C
```{r SFig03,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
layout2ndline <- "
AB
"
plots[["2ndline"]] <- plots[["nGenes_long"]] + plots[["nUMIs_long"]] +  plot_layout(guides = "collect", design = layout2ndline) &  # Collect legends from all plots
  theme(
    legend.position = "right",           
    legend.justification = "center",      
    legend.margin = margin(t = +5),      
    legend.key.size = unit(0.5, "cm"),
    legend.title = element_blank()
  ) &
  plot_annotation(
    tag_levels = list(c('B', " "),  
                      theme = theme(
                        plot.tag = element_text(size = 7, face = "plain")  # Customize tag size and appearance
                      )))

plots[["2ndline"]]

ggsave(
  filename = "results/2V_FigS03_2ndline_boxplot_UMI_detected_genes.svg",              # Name of the file
  plot = plots[["2ndline"]],          # The plot object to save
  device = "svg",                          # Specify the output format (png, pdf, svg)
  width = 16.9,                              # Width of the image
  height = 4.5,                              # Height of the image
  units = "cm"                            # Units for width and height 
)
```

#Violin - replicate - illustrate reads in coding regions or exons; number of genes or UMIs per cell
```{r Violin plots for technical parameters_ALL CELLS  w/o SHAM and MCAO,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_val <- readRDS("ws/1DP_00_val_seurat_SCT_normalised_orders.rds")

DefaultAssay(seurat_val) <- "RNA"

features_list <- list(
  "pct_coding" = "Reads to coding regions",
  "pct_exon" = "Reads to exons",
  "nFeature_RNA" = "Genes/UMI (20K Reads)",
  "nCount_RNA" = "Total number of UMIs"
)

# List for y-axis titles
y_axis_titles <- list(
  "pct_coding" = "Reads in coding regions (%)",
  "pct_exon" = " Reads in exons (%)",
  "nFeature_RNA" = "Number of genes per cell",
  "nCount_RNA" = " Number of UMIs per cell"
)


# Create the violin plot with box plots and custom y-axis titles
for (feature in names(features_list)) {
  plots[[paste0(feature, "_val")]] <- VlnPlot(seurat_val, features = feature, 
                              group.by = "Cond_preparation", 
                              split.by = "Cond_publication_val",
                              pt.size = 0, # Remove individual points
                              cols = col_list[["Conditions_violin_val"]]) +  
    labs(x = "Condition",  y = y_axis_titles[[feature]]) +  # Manually set y-axis title
    theme_mk +
    remove_grid +
    theme(
      strip.background = element_blank(),
       axis.title.x = element_blank(),
      axis.text.x = element_text(angle = 325, vjust = +0.1, hjust = +0.2),
      axis.title.y = element_text(margin = margin(t = -10)),
      plot.title = element_blank()
    ) +
    NoLegend() +
        # Overlay box plot with white fill
    geom_boxplot(width = 0.1, outlier.shape = NA, color = "black", fill = "white")
}

  

layout <- "
AB
CD
"
p <- plots[["pct_coding_val"]] + plots[["pct_exon_val"]] + plots[["nFeature_RNA_val"]] + plots[["nCount_RNA_val"]] +  plot_layout(design =  layout)
p
```

# SFig 01 B
```{r Fig03_2ndline,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
layout_2nd_line <- "
ABCD
"
plots[["violin_wSM_val"]] <- plots[["nFeature_RNA_val"]] + plots[["nCount_RNA_val"]] + plots[["pct_coding_val"]] + plots[["pct_exon_val"]] +  plot_layout(design = layout_2nd_line)  &
  plot_annotation(
    tag_levels = list(c('C', " "),  
                      theme = theme(
                        plot.tag = element_text(size = 7, face = "plain") 
                      )))

plots[["violin_wSM_val"]]

ggsave(
  filename = "results/2V_FigS03_1stline_ViolinPlots_w_ShamMCAOval.svg",             
  plot = plots[["violin_wSM_val"]],         
  device = "svg",                         
  width = 16.9,                              
  height = 4.5,                              
  units = "cm"                             
)
```

#Median values for violin plots 
```{r Median,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Extract metadata from the Seurat object as a data frame
metadata_df <- seurat_val@meta.data

# Create a data frame to store medians for all features
medians_df <- data.frame()

# Loop through the features to calculate medians
for (feature in names(features_list)) {
  
  # Check if the feature exists in the metadata
  if (feature %in% colnames(metadata_df)) {
    
    # Calculate the median for each condition
    feature_medians <- metadata_df %>%
      group_by(Cond_preparation) %>%  # Group by relevant conditions
      summarise(median_value = median(.data[[feature]], na.rm = TRUE)) %>%
      mutate(feature = feature)
    
    # Add the result to the medians_df
    medians_df <- bind_rows(medians_df, feature_medians)
    
  } else {
    message(paste("Feature", feature, "not found in metadata"))
  }
}

# View the extracted medians
print(medians_df)
```