---
title: "2V_SF01_Technical"
author: "Domi"
date: "2024-09-13"
output: html_document
editor_options: 
  chunk_output_type: console
---
## Introduction

This script provides visualizations that demonstrate the presence of debris and doublets—technical artifacts—within  specific cluster.

It is dedicated to visualizing preprocessed data generated from **1DP_01_mg_cells_raw_data_to_seurat_merged_integrated_UMAP.Rmd**. The processed data are visualized using **UMAP plots**, **violin plots**, and **FACS plots** to illustrate clustering and assess data quality.


```{r Load packages and lists, message=FALSE, warning=FALSE, paged.print=TRUE}
suppressMessages(library("Matrix"))
suppressMessages(library("NormExpression"))
suppressMessages(library("DropletUtils"))
suppressMessages(library("pheatmap"))
suppressMessages(library("RColorBrewer"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("sctransform"))
suppressMessages(library("DT"))
suppressMessages(library("scales"))
suppressMessages(library("ggrepel"))
suppressMessages(library("dplyr"))
suppressMessages(library("SeuratWrappers"))
suppressMessages(library("scmap"))
suppressMessages(library("DoubletFinder"))
suppressMessages(library("scCATCH"))
suppressMessages(library("reshape2"))
suppressMessages(library("CellChat"))
suppressMessages(library("ComplexHeatmap"))
suppressMessages(library("here"))
suppressMessages(library("gridExtra"))
suppressMessages(library("clustree"))
suppressMessages(library("magrittr"))
suppressMessages(library("patchwork"))
suppressMessages(library("purrr"))
plots <- list()
col_list <- list()
orders <- list()
source("0F_colours_orders.R", local = F, echo = F, print.eval = F)
source("0F_general_functions.R", local = F, echo = F, print.eval = F)
gene_symbol <- read.table("data/Smartseq3xpress.gene_names.txt", header = TRUE)#File generated with zUMIS
```

#UMAP
```{r UMAP_all_microglia, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_mg <- readRDS("ws/1DP_01_seurat_merged_integrated_UMAP_mg_cells_brm.rds")
DefaultAssay(seurat_mg) <- "SCT"

plots[["UMAP_all_mg_rm"]] <- DimPlot(
  seurat_mg,
  reduction = "umap.cca",
  group.by = "cca_clusters_sbst",
  pt.size = 0.1,
  cols =col_list[["UMAP_mg_rm"]]
) + 
 ggtitle("All conditions") + 
  theme_umap +
  remove_grid +
  remove_axes+
  guides(colour = guide_legend(override.aes = list(size = 1)))  


plots[["UMAP_all_mg_rm"]]
```

```{r UMAP_all_cells_condition_LiveCells, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_filtered_LiveCells <- subset(seurat_mg,  LiveCells != "LiveNuclei + FixedNuclei")

plots[["UMAP_mg_LiveCells_rm"]] <- DimPlot(
  seurat_filtered_LiveCells,
  reduction = "umap.cca",
 group.by = "LiveCells",
  pt.size = 0.1,
  cols = col_list[["UMAPLiveCells"]]
) + 
  ggtitle("LiveCells") + 
 theme_umap +
  remove_grid +
  remove_axes+
  guides(colour = guide_legend(override.aes = list(size = 1)))  


plots[["UMAP_mg_LiveCells_rm"]]
rm(seurat_filtered_LiveCells)
```

```{r UMAP_all_cells_condition_LiveNuclei, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_filtered_LiveNuclei <- subset(seurat_mg, LiveNuclei != "LiveCells + FixedNuclei")


plots[["UMAP_mg_LiveNuclei_rm"]] <- DimPlot(
  seurat_filtered_LiveNuclei,
  reduction = "umap.cca",
  group.by = "LiveNuclei",
  pt.size = 0.1,
  cols = col_list[["UMAPLiveNuclei"]]
) + 
   ggtitle("LiveNuclei") + 
  theme_umap +
  remove_grid +
  remove_axes+
  guides(colour = guide_legend(override.aes = list(size = 1)))

plots[["UMAP_mg_LiveNuclei_rm"]]
rm(seurat_filtered_LiveNuclei)

```

```{r UMAP_all_cells_condition_FixNuclei, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_filtered_FixedNuclei <- subset(seurat_mg, FixedNuclei != "LiveCells + LiveNuclei")


plots[["UMAP_mg_FixedNuclei_rm"]] <- DimPlot(
  seurat_filtered_FixedNuclei,
  reduction = "umap.cca",
  group.by = "FixedNuclei",
  pt.size = 0.1,
  cols = col_list[["UMAPFixedNuclei"]]
) + 
   ggtitle("FixedNuclei") + 
  theme_umap +
  remove_grid +
  remove_axes+
  guides(colour = guide_legend(override.aes = list(size = 1)))  


plots[["UMAP_mg_FixedNuclei_rm"]]
rm(seurat_filtered_FixedNuclei)
```

#SFig 03 A - UMAP visualisation in combination of 4 UMAPs
```{r UMAP saving, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
layout <- "
AB
CD
"
plots[["UMAP_clust_rm"]] <- plots[["UMAP_all_mg_rm"]] + plots[["UMAP_mg_LiveCells_rm"]] + plots[["UMAP_mg_LiveNuclei_rm"]] + plots[["UMAP_mg_FixedNuclei_rm"]] +  plot_layout(design =  layout)&
  plot_annotation(
    tag_levels = list(c('A', " "), 
                      theme = theme(
                        plot.tag = element_text(size = 7, face = "plain") 
                      )))


plots[["UMAP_clust_rm"]]

ggsave(
  filename = "results/2V_FigS02_UMAPs_mg_rm.svg",             
  plot = plots[["UMAP_clust_rm"]],          
  device = "svg",                          
  width = 8,                             
  height = 9,                             
  units = "cm"                            
)

```

#FACS paramters
```{r FACS _ for LiveCellsFCS.AvsSSC.A, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
metadata_df <- read.csv("ws/1DP_01_mg_seurat_metadata_brm.csv")
metadata_df$cca_clusters_sbst <- as.factor(metadata_df$cca_clusters_sbst)

# Remove cells which have values 0 because they weren't measured with index sorting
metadata_df <- metadata_df %>%
  filter(FSC.A != 0)

# Subset for only "LiveCells"
conditions_to_plot <- metadata_df %>%
  subset(Cond_preparation == "LiveCells")

# Create a named vector of colors matching the clusters
cluster_colors <- col_list[["UMAP_mg_rm"]]
names(cluster_colors) <- levels(metadata_df$cca_clusters_sbst)

# Define the pairs of variables to plot
variable_pairs <- list(
  c("FSC.A", "SSC.A"),
  c("FSC.A", "FSC.H")
)


# Assuming conditions_to_plot is a data frame or filtered Seurat object with columns FSC.A, SSC.A
plots[["LiveCellsFCS.AvsSSC.A"]] <- ggplot(conditions_to_plot, aes(x = `FSC.A`, y = `SSC.A`)) +
    geom_point(aes(color = cca_clusters_sbst), size = 0.1) +  
    scale_color_manual(values = cluster_colors) +  
    labs(
      title = "LiveCells",    
      x = "FSC.A",             
      y = "SSC.A",             
      color = "Clusters"     
    ) +
    theme_mk_title +               
    remove_grid +            
    scale_x_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale()), 
                       breaks = scales::pretty_breaks(n = 3)) +  
    scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale()), 
                       limits = c(0, 160000)) +  
    NoLegend()


plots[["LiveCellsFCS.AvsSSC.A"]]

```

```{r FACS _ for LiveCellsFCS.AvsFCS.H, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
plots[["LiveCellsFCS.AvsFSC.H"]] <- ggplot(conditions_to_plot, aes(x = `FSC.A`, y = `FSC.H`)) +
    geom_point(aes(color = cca_clusters_sbst), size = 0.1) +  
    scale_color_manual(values = cluster_colors) + 
    labs(
      title = "LiveCells",    
      x = "FSC.A",             
      y = "FSC.H",             
      color = "Clusters"     
    ) +
    theme_mk_title +               
    remove_grid +           
    scale_x_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale()), 
                       breaks = scales::pretty_breaks(n = 3)) +  
    scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale()), 
                       limits = c(0, 160000)) + 
  NoLegend()


plots[["LiveCellsFCS.AvsFSC.H"]]

```

# SFig 03 B
```{r Fig02 FACS, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#FACS plots for LiveCells
layout <- "
AB
"
# Combine the two plots and bring the legend closer to the charts
plots[["FACS"]] <- (plots[["LiveCellsFCS.AvsSSC.A"]] + 
      plots[["LiveCellsFCS.AvsFSC.H"]]) + 
      plot_layout(guides = "collect") &  # Collect and combine legends
      theme(
        legend.position = "bottom",  
        legend.margin = margin(t = -10),  # Reduce the margin between legend and plots
        plot.margin = margin(10, 10, 10, 10)  # Adjust plot margins to make space for the legend
      ) &  
      guides(colour = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 1)))


# View the combined plot with the legend closer to the charts
plots[["FACS"]]



ggsave(
  filename = "results/2V_FigS02_FACS.svg",              
  plot = plots[["FACS"]],          
  device = "svg",                          
  width = 9.5,                              
  height = 5.5,                             
  units = "cm"                            
)
```


#Violin plots - percentages of ribosomes
```{r Violin_plots percent.ribLiveNUclei, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_live_nuclei <- subset(seurat_mg, Cond_preparation == "LiveNuclei")
Idents(seurat_live_nuclei) <- "cca_clusters_sbst"

plots[["percent.ribLiveNUclei"]] <- VlnPlot(seurat_live_nuclei, features = "percent.rib", 
                              pt.size = 0, # Remove individual points
                              cols = col_list[["UMAP_mg_rm"]],
                              split.by = "cca_clusters_sbst") +  
    labs(title = "LiveCells",
         x = "Clusters",
         y = "Ribosomes (%)"
        ) +
  theme_mk_title +
  remove_grid+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),limits = c(0, 11.5))+ # Start y-axis at 0 with a bit of padding
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(vjust = +2),
    axis.title.y = element_text(margin = margin(t = -10))
    )+
  ggtitle("LiveNuclei") +
  NoLegend()


plots[["percent.ribLiveNUclei"]]

```


```{r Violin_plots percent.ribFixedNUclei, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_fixed_nuclei <- subset(seurat_mg, Cond_preparation == "FixedNuclei")
Idents(seurat_fixed_nuclei) <- "cca_clusters_sbst"


plots[["percent.ribFixedNUclei"]] <- VlnPlot(seurat_fixed_nuclei, features = "percent.rib", 
                              pt.size = 0, # Remove individual points
                              cols = col_list[["UMAP_mg_rm"]],
                              split.by = "cca_clusters_sbst") +  
    labs(x = "Clusters",
         y = NULL
        ) +
  theme_mk_title +
  remove_grid+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), limits = c(0, 11.5))+ # Start y-axis at 0 with a bit of padding
  theme(
    strip.background = element_blank(),
    axis.text.x = element_text(vjust = +2)
    )+
  ggtitle("FixedNuclei") +
  NoLegend()


plots[["percent.ribFixedNUclei"]]
```

#SFig 03CD
```{r Fig02 violin, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
layout <- "
AB
"
plots[["violin"]] <- plots[["percent.ribLiveNUclei"]] + plots[["percent.ribFixedNUclei"]] +  plot_layout(design =  layout)&
  plot_annotation(
    tag_levels = list(c('C', 'D'),  # Use 'a' for automatic labels ('a', 'b', 'c', 'd')
                      theme = theme(
                        plot.tag = element_text(size = 7, face = "plain")  # Customize tag size and appearance
                      )))
plots[["violin"]]

ggsave(
  filename = "results/2V_FigS02_violin.svg",              
  plot = plots[["violin"]],          
  device = "svg",                          
  width = 9,                          
  height = 4.5,                            
  units = "cm"                           
)
```
