---
title: "2V_00_all_cells_UMAP_dot_plot_bar_plot"
author: "Dominika Dostalova"
date: "15 01 2025"
output: html_document
editor_options: 
  chunk_output_type: console
---
Introduction
This script is dedicated to visualizing preprocessed data from **1DP_00_all_cells_raw_data_to_seurat_merged_integrated_UMAP.Rmd**. The processed data are represented using **UMAP plots, dot plots, and stacked plots** to illustrate clustering, marker expression, and population distributions. These visualizations provide insights into cell-type composition and gene expression patterns across different isolation protocols.


Loading required libraries and functions
```{r Load packages and lists,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
suppressMessages(library("Matrix"))
suppressMessages(library("RColorBrewer"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("sctransform"))
suppressMessages(library("scales"))
suppressMessages(library("dplyr"))
suppressMessages(library("SeuratWrappers"))
suppressMessages(library("reshape2"))
suppressMessages(library("ComplexHeatmap"))
suppressMessages(library("gridExtra"))
suppressMessages(library("magrittr"))
suppressMessages(library("patchwork"))
plots <- list()
col_list <- list()
orders <- list()
source("0F_general_functions.R", local = F, echo = F, print.eval = F)
source("0F_colours_orders.R", local = F, echo = F, print.eval = F)
#Read table with gene symbols end IDs
gene_symbol <- read.table("data/Smartseq3xpress.gene_names.txt", header = TRUE)
```

```{r Load seurat object,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat <- readRDS("ws/1DP_00_seurat_merged_integrated_UMAP_all_cells.rds")
```

UMAP visualisation
```{r UMAP_all_cells,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
plots[["UMAP_all"]] <- DimPlot(
  seurat,
  reduction = "umap.cca",
  group.by = "Cell_types_all_abbr",
  pt.size = 0.1,
  cols = col_list[["UMAP_all"]]
) + 
  ggtitle("All conditions") + 
  theme_umap +
  remove_grid +
  remove_axes+
  guides(colour = guide_legend(override.aes = list(size = 1)))  


plots[["UMAP_all"]]
```

```{r UMAP_all_cells_condition_LiveCells,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_filtered_LiveCells <- subset(seurat, LiveCells != "LiveNuclei + FixedNuclei")


plots[["UMAP_all_LiveCells"]] <- DimPlot(
  seurat_filtered_LiveCells,
  reduction = "umap.cca",
  group.by = "LiveCells",
  pt.size = 0.1,
  cols = col_list[["UMAPLiveCells"]]
) + 
  ggtitle("LiveCells") + 
 theme_umap +
  remove_grid +
  remove_axes+
  guides(colour = guide_legend(override.aes = list(size = 1))) 

plots[["UMAP_all_LiveCells"]]
#rm(seurat_filtered_LiveCells)
```

```{r UMAP_all_cells_condition_LiveNuclei,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_filtered_LiveNuclei <- subset(seurat, LiveNuclei != "LiveCells + FixedNuclei")

plots[["UMAP_all_LiveNuclei"]] <- DimPlot(
  seurat_filtered_LiveNuclei,
  reduction = "umap.cca",
  group.by = "LiveNuclei",
  pt.size = 0.1,
  cols = col_list[["UMAPLiveNuclei"]]
) + 
  ggtitle("LiveNuclei") + 
  theme_umap +
  remove_grid +
  remove_axes+
  guides(colour = guide_legend(override.aes = list(size = 1)))  

plots[["UMAP_all_LiveNuclei"]]
#rm(seurat_filtered_LiveNuclei)
```

```{r UMAP_all_cells_condition_FixedNuclei,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_filtered_FixedNuclei <- subset(seurat, FixedNuclei != "LiveCells + LiveNuclei")

plots[["UMAP_all_FixedNuclei"]] <- DimPlot(
  seurat_filtered_FixedNuclei,
  reduction = "umap.cca",
  group.by = "FixedNuclei",
  pt.size = 0.1,
  cols = col_list[["UMAPFixedNuclei"]]
) + 
  ggtitle("FixedNuclei") + 
  theme_umap +
  remove_grid +
  remove_axes+
  guides(colour = guide_legend(override.aes = list(size = 1)))  

plots[["UMAP_all_FixedNuclei"]]
#rm(seurat_filtered_FixedNuclei)
```

```{r Table with cells per condition, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#info about cells per condition
table(seurat@meta.data$Cond_preparation)
table(seurat@meta.data$Cell_types_all_abbr)
```

## Fig. 01 B - UMAP visualisation in combination of 4 UMAPs
```{r Save UMAP visualisation, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
layout_1stline <- "
AB
CD
"
plots[["1st_line01"]] <- plots[["UMAP_all"]] + plots[["UMAP_all_LiveCells"]] +
plots[["UMAP_all_LiveNuclei"]] + plots[["UMAP_all_FixedNuclei"]] +  
   plot_layout(design =  layout_1stline,
              widths = c(1, 1)) +
  plot_annotation(
    tag_levels = list(c('B'),  
                      theme = theme(
                        plot.tag = element_text(size = 6, face = "plain")  
                      )))

plots[["1st_line01"]]

ggsave(
  filename = "results/2V_Fig01_1stline.svg",              
  plot = plots[["1st_line01"]],          
  device = "svg",                          
  width = 9.3,                             
  height = 9,                              
  units = "cm"                            
)
```


Dotplot visualisation of genes
```{r List of marker_genes_all, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#makrers chosen by web site https://cellxgene.cziscience.com
markers_all <- c("Hexb", "Cx3cr1", "Siglech", "Trem2",  #Microglia
                 "Lyz2","Pf4", "Ms4a7", "F13a1",        #PVM - mousebrain.org
                 "Ms4a6c", "Tmsb10", "S100a4","Ccr2",   #Monocytes
                 "S100a8", "S100a9","Mmp9", "Msrb1",    #Neutrophils
                 "Cald1", "Vtn",   "Myl9", "Slc38a11",  #Mural cells
                 "Egfl7", "Eng", "Ramp2", "Cd93",       #Endothelial cells
                 "Mbp", "Plp1", "Enpp2", "Mag",         #Oligo
                 "Slc1a2", "Prdx6", "Mt1","Slc4a4")     #Astrocytes

```

```{r RNA normalization, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
DefaultAssay(seurat) <- "RNA" 
seurat <- JoinLayers(seurat, assay = "RNA")
all.genes.RNA <- rownames(seurat)
seurat <- NormalizeData(seurat, normalization.method = "LogNormalize", scale.factor = 10000)
seurat <- ScaleData(seurat, features = all.genes.RNA)
```

```{r Changed Dot plot with feature test, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
DefaultAssay(seurat) <- "RNA" 
Idents(seurat) <- "Cell_types_all_abbr"

plots[["dot_plot_all"]] <- DotPlot(
  seurat, 
  features = Symbol_to_ENSEMBL(markers_all), 
 dot.min = 0.20 
) + 
  scale_x_discrete(labels = ENSEMBL_to_Symbol(Symbol_to_ENSEMBL(markers_all))) + 
  theme_mk + 
  remove_grid +
  theme(
    axis.text.x = element_text(angle = 290, hjust = 0, vjust = 0.5, size = 6, face = "italic"), 
    axis.title.x = element_blank(),                                
    axis.title.y = element_blank(),                                
    legend.position = "bottom",                                   
    plot.margin = margin(0,0,0,0),                                 
    
  legend.margin = margin(b = -5, l = -3),
  legend.box.spacing = unit(0.1, "cm"),
  legend.box.just = "left"
    ) +
  scale_color_gradientn(
    name = "Scaled average \ngene expression",   
    colors = c("#E8ECEF","#9ABDDC", "#3A0956")            
  ) +
  scale_size_continuous(
    name = "Pergentage \nof cells (%)",
    range = c(0.5, 2)              
  ) +
  guides(
    color = guide_colorbar(title.position = "left"), 
    size = guide_legend(title.position = "left")     
  )

# Display the customized Dot Plot
plots[["dot_plot_all"]]
```

## Fig. 01 C
```{r Fig01_2nd line DotPlot, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
layout_2ndline <- "
A
"
plots[["2nd_line_dot01"]] <- plots[["dot_plot_all"]] +
  plot_layout(design =  layout_2ndline) +
  plot_annotation(
    tag_levels = list(c('C'), 
                      theme = theme(
                        plot.tag = element_text(size = 7, face = "plain")  
                      )))
                      
plots[["2nd_line_dot01"]]

ggsave(
  filename = "results/2V_Fig01_2ndline_dot.svg",          
  plot = plots[["2nd_line_dot01"]],          
  device = "svg",                          
  width = 9.35,                             
  height = 5,                              
  units = "cm"                          
)
```


Stacked bar_plots with relative proportion of cells in clusters
```{r Counts for stacked bar plot, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
metadata_df <- read.csv("ws/1DP_00_seurat_all_cells_metadata_for_subset_mg.csv")


#Filter NA
metadata_df <- metadata_df %>%
  filter(!is.na(Cell_types_all_abbr))


# Calculate total counts per condition
total_counts_per_condition <- metadata_df %>%
  group_by(Cond_publication) %>%
  tally(name = "total_cells")

# Calculate counts per cluster per condition
cluster_counts <- metadata_df %>%
  group_by(Cond_publication, Cell_types_all_abbr) %>%
  tally(name = "cluster_count")

# Join the total counts with cluster counts
perc_clust_per_condition <- cluster_counts %>%
  left_join(total_counts_per_condition, by = "Cond_publication") %>%
  mutate(percent = (cluster_count / total_cells) * 100) %>%
  select(Cond_publication, Cell_types_all_abbr, percent)

# View the resulting data frame
print(perc_clust_per_condition)

#Check microglia 
perc_clust_per_condition %>%
  filter(Cell_types_all_abbr == "MG")

rm(total_counts_per_condition, cluster_counts)

#change order of cell types - inverted
perc_clust_per_condition$Cell_types_all_abbr <- factor(perc_clust_per_condition$Cell_types_all_abbr, levels = orders[["Cell_types_all_abbr_invert"]])

#group Sham together and MCAO 
perc_clust_per_condition$Cond_publication <- factor(perc_clust_per_condition$Cond_publication, levels = orders[["Cond_publication"]])



# Group microglia cluster into one - better for visualisation 
perc_clust_per_condition %<>% mutate("Cond_SHAMvsMCAO" = factor(case_when(
  Cond_publication == "LiveCells_Sham" ~ "Sham",
  Cond_publication == "LiveNuclei_Sham"~ "Sham",
  Cond_publication == "FixedNuclei_Sham" ~ "Sham",
  Cond_publication ==  "LiveCells_MCAO"~ "MCAO",
  Cond_publication == "LiveNuclei_MCAO" ~ "MCAO",
  Cond_publication == "FixedNuclei_MCAO" ~ "MCAO"
)))

perc_clust_per_condition$Cond_SHAMvsMCAO <- factor(perc_clust_per_condition$Cond_SHAMvsMCAO , levels = orders[["Cond_SHAMvsMCAO"]])
perc_clust_per_condition$Cond_SHAMvsMCAO %>% table()

perc_clust_per_condition %<>% mutate("Cond_appr" = factor(case_when(
  Cond_publication == "LiveCells_Sham" ~ "LiveCells",
  Cond_publication == "LiveNuclei_Sham"~ "LiveNuclei",
  Cond_publication == "FixedNuclei_Sham" ~ "FixedNuclei",
  Cond_publication ==  "LiveCells_MCAO"~ "LiveCells",
  Cond_publication == "LiveNuclei_MCAO" ~ "LiveNuclei",
  Cond_publication == "FixedNuclei_MCAO" ~ "FixedNuclei"
)))
```

```{r Stacked bar plot - condition vs percentage of cells, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
plots[["stacked_bar_all"]] <- ggplot(perc_clust_per_condition, aes(x = Cond_appr, y = percent, fill = Cell_types_all_abbr)) +
  geom_bar(position = "stack", stat = "identity", color = "#2d1800", size = 0.3 ) + 
  scale_fill_manual(values = col_list[["UMAP_all"]]) +  
  labs(x = "Conditions", y = "Relative proportion of cells (%)", fill = "Cell \ntypes") +
 theme_mk +
   scale_y_continuous(
    expand = c(0, 0),               
    breaks = seq(0, 100, by = 20))+ 
  facet_grid(. ~ Cond_SHAMvsMCAO)+
  remove_grid +
   theme(
    strip.text = element_text(size = 6),        
    strip.background = element_blank(),        
    legend.position = "right",
    legend.margin = margin(t = -7.5),
    plot.margin = margin(0,0,0,0),
    axis.text.x = element_text(angle = 310, vjust = +0.1, hjust = +0.2),
    axis.title.x = element_blank(), 
    legend.text = element_text(margin = margin(t = 0, r = 1, b = 1, l = 1)),
    legend.title = element_text(margin = margin(t = 0, r = 1, b = 6, l = 1))
    )  

plots[["stacked_bar_all"]]
```

# Fig. 01 D
```{r Save Stacked bar plot,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
ggsave(
  filename = "results/2V_Fig01_2ndline_stacked_barplot.svg",              
  plot = plots[["stacked_bar_all"]],          
  device = "svg",                          
  width = 7,                             
  height = 4.8,                              
  units = "cm"                          
)
```