---
title: "2V_04_DEG"
author: "Dominika Dostalova"
date: "2024-11-27"
output: html_document
editor_options: 
  chunk_output_type: console
---
Introduction
This script is dedicated to visualizing processed data from **1DP_02_DEG.Rmd**. The processed data are represented using **Volcano and Scatter plots** to illustrate DEG analysis.

#Loading of improtant files
```{r Load packages,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
suppressMessages(library("Matrix"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("sctransform"))
suppressMessages(library("ggrepel"))
suppressMessages(library("dplyr"))
suppressMessages(library("magrittr"))
suppressMessages(library("patchwork"))
suppressMessages(library("scales"))
suppressMessages(library("tidyr"))
suppressMessages(library("openxlsx"))
plots <- list()
col_list <- list()
orders <- list()
source("0F_colours_orders.R", local = F, echo = F, print.eval = F)
source("0F_general_functions.R", local = F, echo = F, print.eval = F)
gene_symbol <- read.table("data/Smartseq3xpress.gene_names.txt", header = TRUE)#File generaetd with zUMIS
```

#Volcano Plots for comparison DEG in HM and ARM across isolation protocols
```{r Volcano,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Load the combined RDS file containing all conditions
de_results_all <- readRDS("ws/1DP_02_DEG_ARMvsHM_allconditions.rds")

# Define genes of interest - ARM Frigerio2019
genes_of_interest <- c(
  "Apoe", "Cst7", "Lpl", "Clec7a", "Lyz2", "Ccl3", "Lgals3bp", "Cd63", "H2-D1",  
  "Ctsb", "Cd9", "Ctsl", "Fth1", "Ctsz", "Cd52", "Ctsd", "B2m", "Eef1a1", "Cd83", 
  "Ccl6", "Hif1a", "Tyrobp", "Serpine2", "Cadm1", "H2-K1", "Npc2", "Pkm", 
  "Aldoa", "Tpi1", "Pld3", "Gusb", "Plek", "Ldha"
)

# Define protocol-specific colors (ordered for use with list indices)
protocol_colors <- c("#5F7048", "#8a5c1a", "#d9741a")  # Dark Green, Brown, Orange

# Get protocol names
protocol_names <- names(de_results_all)

# Initialize an empty list to store plots
plots <- list()

# Loop through each protocol
for (i in seq_along(protocol_names)) {
  protocol <- protocol_names[i]
  
  # Extract DE results
  de_results <- de_results_all[[protocol]]
  
  # Map ENSEMBL IDs to gene symbols
  de_results$gene_symbol <- gene_symbol$gene_name[match(rownames(de_results), gene_symbol$gene_id)]
  
  # Add significance
  de_results$significant <- with(de_results, p_val_adj < 0.05 & abs(avg_log2FC) > 1)
  
  # Add coloring category
  de_results$color_group <- ifelse(
    de_results$avg_log2FC > -1 & de_results$avg_log2FC < 1,
    "lowFC",
    ifelse(de_results$significant, "significant", "nonsig")
  )
  
  # Filter genes of interest
  gene_inter <- de_results %>%
    filter(gene_symbol %in% genes_of_interest)
  
  # Volcano plot
  plots[[protocol]] <- ggplot(data = de_results,
                              aes(x = avg_log2FC, y = -log10(p_val_adj))) +
    geom_point(aes(colour = color_group),
               alpha = 0.6,
               shape = 16,
               size = 1) +
    geom_point(data = gene_inter,
               aes(x = avg_log2FC, y = -log10(p_val_adj)),
               shape = 21,
               fill = "black",
               color = "black",
               size = 1,
               stroke = 0.2) +
   geom_label_repel(data = gene_inter %>% filter(significant),
                     aes(label = gene_symbol),
                      box.padding   = 0.22,
                     label.padding = 0.10,
                     point.padding = 0.1,
                     color = "black",
                     size = 1.8,
                     max.iter = 20000,
                     max.overlaps = 20,
                     min.segment.length = 0,
                     force = 30,
                     nudge_y = 1,
                     label.size = 0,
                     segment.size = 0.1) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "#B6C7DB", linewidth = 0.1) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "#B6C7DB", linewidth = 0.1) +
    labs(
      title = paste(protocol, "\n(ARM vs HM)"),
      subtitle = ("   down in ARM                          up in ARM"),
      x = "Log2 fold change",
      y = if (i == 1) "-Log10 adjusted p-value" else NULL
    ) +
    scale_color_manual(
      values = c(
        "nonsig" = "gray80",
        "lowFC" = "gray80",
        "significant" = protocol_colors[i]
      ),
      guide = "none"
    ) +
    scale_x_continuous(limits = c(-6.5, 9.1)) +
    scale_y_continuous(limits = c(-5, NA )) +#70
    theme_mk_title +
    remove_grid
  
  print(paste("Volcano plot created for:", protocol))
}

# Combine plots for viewing
plots[["volcano"]] <- plots[["LiveCells"]] + plots[["LiveNuclei"]] + plots[["FixedNuclei"]]
plots[["volcano"]]
```

# Fig 03 A
```{r Fig03_1stline,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
layout_1stline <- "
ABC
"
plots[["1st_line"]] <-  plots[["LiveCells"]] + plots[["LiveNuclei"]] + plots[["FixedNuclei"]] +
  plot_layout(guides = "collect", design = layout_1stline) &  
  theme(
    legend.position = "bottom",           
    legend.justification = "center",      
    legend.margin = margin(t = -10)       
  ) &
  plot_annotation(
    tag_levels = list(c('A', " "),  
                      theme = theme(
                        plot.tag = element_text(size = 7, face = "plain")  
                      )))

plots[["1st_line"]]


ggsave(
  filename = "results/2V_Fig03_1stline_VolcanoPlots.svg",              
  plot = plots[["1st_line"]],          
  device = "svg",                         
  width = 16.9,                              
  height = 7,                              
  units = "cm"                           
)
```


#SCATTER PLOTS - differences in mean expression of genes in ARM population
#Scatter plot LiveCells vs FixedNuclei in ARM
```{r Table for ScatterPlot LiveCells vs FixedNuclei,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#gene_table - helper for faster dividing of ARM genes into slots for ScatterPlot
gene_table <- readRDS("ws/1DP_03_gene_table_ScatterPlot_LCvsFN.rds")

list_gene_names_ARM_sym <- c(
  "Apoe", "Cst7", "Lpl", "Clec7a", "Lyz2", "Ccl3", "Lgals3bp", "Cd63", "H2-D1",  
  "Ctsb", "Cd9", "Ctsl", "Fth1", "Ctsz", "Cd52", "Ctsd", "B2m", "Eef1a1", "Cd83", 
  "Ccl6", "Hif1a", "Tyrobp", "Serpine2", "Cadm1", "H2-K1", "Npc2", "Pkm", 
  "Aldoa", "Tpi1", "Pld3", "Gusb", "Plek", "Ldha"
)


# Assuming `gene_table` is your data frame
# Initialize vectors for each category
up2_fn <- gene_table$`Upregulated >2x in FN`[gene_table$`Upregulated >2x in FN` %in% list_gene_names_ARM_sym]
up1_fn <- gene_table$`Upregulated 1-2x in FN`[gene_table$`Upregulated 1-2x in FN` %in% list_gene_names_ARM_sym]
up0_fn <- gene_table$`Upregulated 0-1x in FN`[gene_table$`Upregulated 0-1x in FN` %in% list_gene_names_ARM_sym]
up2_lc <- gene_table$`Upregulated >2x in LC`[gene_table$`Upregulated >2x in LC` %in% list_gene_names_ARM_sym]
up1_lc <- gene_table$`Upregulated 1-2x in LC`[gene_table$`Upregulated 1-2x in LC` %in% list_gene_names_ARM_sym]
up0_lc <- gene_table$`Upregulated 0-1x in LC`[gene_table$`Upregulated 0-1x in LC` %in% list_gene_names_ARM_sym]

# Combine results into a named list for clarity
result <- list(
  "Upregulated >2x in FN" = up2_fn,
  "Upregulated 1-2x in FN" = up1_fn,
  "Upregulated 0-1x in FN" = up0_fn,
  "Upregulated >2x in LC" = up2_lc,
  "Upregulated 1-2x in LC" = up1_lc,
  "Upregulated 0-1x in LC" = up0_lc
)

# Print results
print(result)
```

```{r Scatter plot LiveCells vs FixedNuclei,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
topT <- readRDS("ws/1DP_03_top_genes_ARM_file.rds")

# Define gene groups with their specific colors
gene_groups <- list(
  up2 = list(genes = Symbol_to_ENSEMBL(c("Fth1", "Eef1a1")), color = "#5B1F3D"),
  up1 = list(genes = Symbol_to_ENSEMBL(c("Cd52", "Apoe", "Ctsd", "Ctsz", "Ccl6", "Npc2", "Ctsl", "Ctsb", "Tpi1", "Cd63", "Cd9", "Tyrobp", "Aldoa","Pkm", "B2m", "Ldha", "Cst7", "Lyz2", "H2-D1")), color = "#A7586D"),
  up0 = list(genes = Symbol_to_ENSEMBL(c("Ccl3",  "Pld3", "Cd83", "Lpl", "Plek",  "Gusb", "Serpine2", "Lgals3bp", "H2-K1", "Clec7a")), color = "#8aa5bf")
)

# Add a `group` and `label_color` column to `topT`
topT$group <- NA
topT$label_color <- NA
for (group in names(gene_groups)) {
  genes <- toupper(gene_groups[[group]]$genes)
  topT$group[toupper(rownames(topT)) %in% genes] <- group
  topT$label_color[toupper(rownames(topT)) %in% genes] <- gene_groups[[group]]$color
}


plots[["scatter_plotLCvsFN"]] <- ggplot(topT, aes(x = LiveCells, y = FixedNuclei)) +
  labs(
    x = "Log normalized mean expression - LiveCells",
    y = "Log normalized mean expression - FixedNuclei"
  ) +
  geom_abline(slope = 1, intercept = c(0, -2, -1, 1, 2), linetype = c(1, 3, 3, 3, 3)) +
  
  # Base points with color based on differences
  geom_point(
    aes(color = ifelse(
      LiveCells - FixedNuclei > 1, "#5F7048",      # Strongly up in LiveCells
      ifelse(
        FixedNuclei - LiveCells > 1, "#d9741a",       # Strongly up in FixedNuclei
        "#A1B8D4"                                 # Similar expression
      )
    )),
    size = 0.5,
    na.rm = TRUE
  ) +
  
  # Highlight gene groups
  geom_point(data = subset(topT, !is.na(group)), aes(color = group), size = 0.5) +
  

  # Add gene labels only for up1 and up2
  geom_label_repel(
    data = subset(topT, group %in% c("up1", "up2")),
    aes(
      label = ENSEMBL_to_Symbol(rownames(subset(topT, group %in% c("up1", "up2")))),
      color = label_color
    ),
    box.padding = 0.22,
    label.padding = 0.1,
    point.padding = 0.1,
    size = 1.8,
    max.iter = 20000,
    min.segment.length = 0,
    max.overlaps = 20,
    show.legend = FALSE,
    label.size = 0, 
    segment.size = 0.1
  )+
  
  # Update the manual color scale with more colors
  scale_color_manual(
    values = c(
      "black",    # labels >2 LiveCells
      "#5F7048",       # points >1 LiveCells
     # "#7286AA",    # labels for genes <1
      "gray80",    # points < 1
      "black",       # labels >1 LiveCells
      "#d9741a", # points >1 FixedNuclei
      "black",    # points for genes <1
      "black",       # points for genes >1 LiveCells
      "black"   # points for genes >2 LiveCells
    )
  ) +
  theme_mk +
  remove_grid +
  theme(
    legend.position = "none"  
  )

plots[["scatter_plotLCvsFN"]]
```

#Scatter plot LiveNuclei vs FixedNuclei in ARM
```{r Table for ScatterPlot LiveNuclei vs FixedNuclei,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#gene_table - helper for faster dividing of ARM genes into slots for ScatterPlot
gene_table <- readRDS("ws/1DP_03_gene_table_ScatterPlot_LNvsFN.rds")

#filtering for ARM genes

list_gene_names_ARM_sym <- c(
  "Apoe", "Cst7", "Lpl", "Clec7a", "Lyz2", "Ccl3", "Lgals3bp", "Cd63", "H2-D1",  
  "Ctsb", "Cd9", "Ctsl", "Fth1", "Ctsz", "Cd52", "Ctsd", "B2m", "Eef1a1", "Cd83", 
  "Ccl6", "Hif1a", "Tyrobp", "Serpine2", "Cadm1", "H2-K1", "Npc2", "Pkm", 
  "Aldoa", "Tpi1", "Pld3", "Gusb", "Plek", "Ldha"
)

# Assuming `gene_table` is your data frame
# Initialize vectors for each category
up2_fn <- gene_table$`Upregulated >2x in FN`[gene_table$`Upregulated >2x in FN` %in% list_gene_names_ARM_sym]
up1_fn <- gene_table$`Upregulated 1-2x in FN`[gene_table$`Upregulated 1-2x in FN` %in% list_gene_names_ARM_sym]
up0_fn <- gene_table$`Upregulated 0-1x in FN`[gene_table$`Upregulated 0-1x in FN` %in% list_gene_names_ARM_sym]
up2_ln <- gene_table$`Upregulated >2x in LN`[gene_table$`Upregulated >2x in LN` %in% list_gene_names_ARM_sym]
up1_ln <- gene_table$`Upregulated 1-2x in LN`[gene_table$`Upregulated 1-2x in LN` %in% list_gene_names_ARM_sym]
up0_ln <- gene_table$`Upregulated 0-1x in LN`[gene_table$`Upregulated 0-1x in LN` %in% list_gene_names_ARM_sym]

# Combine results into a named list for clarity
result <- list(
  "Upregulated >2x in FN" = up2_fn,
  "Upregulated 1-2x in FN" = up1_fn,
  "Upregulated 0-1x in FN" = up0_fn,
  "Upregulated >2x in LN" = up2_ln,
  "Upregulated 1-2x in LN" = up1_ln,
  "Upregulated 0-1x in LN" = up0_ln
)

# Print results
print(result)
```

```{r Scatter plot LiveNuclei vs FixedNuclei,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Load gene expression table for LN vs FN
topT_LNvsFN <- readRDS("ws/1DP_03_top_genes_ARM_file.rds")

# Define gene groups with their specific colors
gene_groups_LNvsFN <- list(
  up1 = list(genes = Symbol_to_ENSEMBL(c("Lyz2")), color = "#A7586D"),
  up0 = list(genes = Symbol_to_ENSEMBL(c(
    "Cd83", "Ctsz", "Ccl6", "Hif1a", "Ctsl", "Serpine2", "Cd9", "Lgals3bp", 
    "B2m", "H2-K1", "Cst7", "H2-D1", "Clec7a", "Cd52", "Ccl3", "Apoe",  
    "Pld3", "Ctsd", "Lpl", "Plek", "Npc2", "Ctsb", "Tpi1", "Fth1",  
    "Cd63", "Gusb", "Tyrobp", "Aldoa", "Cadm1", "Pkm", "Eef1a1", "Ldha"
  )), color = "#8aa5bf")
)

# Add `group` and `label_color` to a copy of topT
topT_LNvsFN$group <- NA
topT_LNvsFN$label_color <- NA
for (group in names(gene_groups_LNvsFN)) {
  genes <- toupper(gene_groups_LNvsFN[[group]]$genes)
  topT_LNvsFN$group[toupper(rownames(topT_LNvsFN)) %in% genes] <- group
  topT_LNvsFN$label_color[toupper(rownames(topT_LNvsFN)) %in% genes] <- gene_groups_LNvsFN[[group]]$color
}

# Create the plot
plots[["scatter_plotLNvsFN"]] <- ggplot(topT_LNvsFN, aes(x = LiveNuclei, y = FixedNuclei)) +
  labs(
    x = "Log normalized mean expression - LiveNuclei",
    y = "Log normalized mean expression - FixedNuclei"
  ) +
  geom_abline(slope = 1, intercept = c(0, -2, -1, 1, 2), linetype = c(1, 3, 3, 3, 3)) +

  # Base points
  geom_point(
    aes(color = ifelse(
      LiveNuclei - FixedNuclei > 1, "#8a5c1a",
      ifelse(FixedNuclei - LiveNuclei > 1, "#d9741a", "#A1B8D4")
    )),
    size = 0.5,
    na.rm = TRUE
  ) +

  # Highlight gene groups
  geom_point(data = subset(topT_LNvsFN, !is.na(group)), aes(color = group), size = 0.5) +
  
   # Add labels only for up1
  geom_label_repel(
    data = subset(topT_LNvsFN, group %in% c("up1")),
    aes(
      label = ENSEMBL_to_Symbol(rownames(subset(topT_LNvsFN, group %in% c("up1")))),
      color = label_color
    ),
    box.padding = 0.22,
    label.padding = 0.1,
    point.padding = 0.1,
    size = 1.8,
    max.iter = 20000,
    min.segment.length = 0,
    max.overlaps = 20,
    show.legend = FALSE,
    label.size = 0,
    segment.size = 0.1
  ) +
  
  # Update the manual color scale with more colors
  scale_color_manual(
    values = c(
      "#8a5c1a",    # points >1 LiveNuclei
      #"#7286AA",       #  labels <1 
      "gray80",    #  points < 1
      "black",    # labels > LiveNuclei
      "#d9741a",       #points >1 FixedNuclei
      "black", # points for genes <1
      "black"    # points >1 LiveNuclei SF
    ),
    guide = "none"
  ) +
  theme_mk +
  remove_grid +
  theme(
    legend.position = "none",  # Show legend for color mapping
  )

plots[["scatter_plotLNvsFN"]]
```

# Fig 03 C
```{r Fig03_1stline,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
layout_1stline <- "
AB"

plots[["1st_line"]] <- plots[["scatter_plotLCvsFN"]] + plots[["scatter_plotLNvsFN"]]+
     plot_layout(design = layout_1stline)

plots[["1st_line"]]

ggsave(
  filename = "results/2V_Fig03_1stline_ScatterPlotLCvsFNt.svg",              # Name of the file
  plot = plots[["1st_line"]],          # The plot object to save
  device = "svg",                          # Specify the output format (png, pdf, svg)
  width = 12,                              # Width of the image
  height = 6,                              # Height of the image
  units = "cm"                            # Units for width and height 
)
```


#Venn diagram with ARM activation - extraction only values
```{r Table with names from DEG,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Create an empty list to store the filtered results
filtered_de_results <- list()

# Create an empty list to store filtered genes for each condition
filtered_genes <- list()

# Iterate over the differential expression results stored in all_de_results
for (name in names(de_results_all)) {
  # Get the DE results for the current iteration
  de_results <- de_results_all[[name]]
  
  # Filter the results based on the condition: avg_log2FC > 1
  filtered_results <- subset(de_results, p_val_adj < 0.05 & avg_log2FC > 1)
  
  # Save the filtered results into the new list
  filtered_de_results[[name]] <- filtered_results
  
  # Store only the gene names that meet the condition avg_log2FC > 1
  filtered_genes[[name]] <- rownames(filtered_results)
  
  # Print a message for tracking progress
  print(paste("Filtered DE results for:", name))
}

# Create the combined table with gene presence/absence
gene_overlap_table <- data.frame(
  Gene = unique(unlist(filtered_genes)),
  LiveCells = unique(unlist(filtered_genes)) %in% filtered_genes$LiveCells,
  LiveNuclei = unique(unlist(filtered_genes)) %in% filtered_genes$LiveNuclei,
  FixedNuclei = unique(unlist(filtered_genes)) %in% filtered_genes$FixedNuclei
)

# Add overlap and status information
gene_overlap_table <- gene_overlap_table %>%
  rowwise() %>%
  mutate(
    OverlapCount = sum(c_across(LiveCells:FixedNuclei)),
    Status = case_when(
      OverlapCount == 3 ~ "Overlap All",
      OverlapCount == 2 & LiveCells & LiveNuclei ~ "LiveCells and LiveNuclei",
      OverlapCount == 2 & LiveCells & FixedNuclei ~ "LiveCells and FixedNuclei",
      OverlapCount == 2 & LiveNuclei & FixedNuclei ~ "LiveNuclei and FixedNuclei",
      OverlapCount == 1 & LiveCells ~ "Unique LiveCells",
      OverlapCount == 1 & LiveNuclei ~ "Unique LiveNuclei",
      OverlapCount == 1 & FixedNuclei ~ "Unique FixedNuclei",
      TRUE ~ "Other"
    )
  )

# Split genes by category
genes_by_category <- gene_overlap_table %>%
  select(Gene, Status) %>%
  group_by(Status) %>%
  mutate(GeneIndex = row_number()) %>% # Create an index for each gene in a category
  pivot_wider(names_from = Status, values_from = Gene, values_fill = NA)


# Save the horizontal table to an Excel file
wb <- createWorkbook()
addWorksheet(wb, "Gene Overlap Analysis")
writeData(wb, "Gene Overlap Analysis", genes_by_category)
saveWorkbook(wb, "ws/2V_04_DEG_gene_overlap.xlsx", overwrite = TRUE)
```

Gene numbers for DEG Venn diagram
```{r Venn,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Create an empty list to store the filtered results
filtered_de_results <- list()

# Create an empty list to store filtered genes for each condition
filtered_genes <- list()

# Iterate over the differential expression results stored in all_de_results
for (name in names(de_results_all)) {
  # Get the DE results for the current iteration
  de_results <- de_results_all[[name]]
  
  # Filter the results based on the condition: avg_log2FC > 1
  filtered_results <- subset(de_results, p_val_adj < 0.05 & avg_log2FC > 1)
  
  # Save the filtered results into the new list
  filtered_de_results[[name]] <- filtered_results
  
  # Store only the gene names that meet the condition avg_log2FC > 1
  filtered_genes[[name]] <- rownames(filtered_results)
  
  # Print a message for tracking progress
  print(paste("Filtered DE results for:", name))
}


# Create a combined table with presence/absence of genes in each condition
gene_overlap_table <- data.frame(
  Gene = unique(unlist(filtered_genes)),
  LiveCells = unique(unlist(filtered_genes)) %in% filtered_genes$LiveCells,
  LiveNuclei = unique(unlist(filtered_genes)) %in% filtered_genes$LiveNuclei,
  FixedNuclei = unique(unlist(filtered_genes)) %in% filtered_genes$FixedNuclei
)

# Add overlap and status information
gene_overlap_table <- gene_overlap_table %>%
  rowwise() %>%
  mutate(
    OverlapCount = sum(c_across(LiveCells:FixedNuclei)),
    Status = case_when(
      OverlapCount == 3 ~ "Overlap All",
      OverlapCount == 2 & LiveCells & LiveNuclei ~ "LiveCells and LiveNuclei",
      OverlapCount == 2 & LiveCells & FixedNuclei ~ "LiveCells and FixedNuclei",
      OverlapCount == 2 & LiveNuclei & FixedNuclei ~ "LiveNuclei and FixedNuclei",
      OverlapCount == 1 & LiveCells ~ "Unique LiveCells",
      OverlapCount == 1 & LiveNuclei ~ "Unique LiveNuclei",
      OverlapCount == 1 & FixedNuclei ~ "Unique FixedNuclei",
      TRUE ~ "Other"
    )
  )

# Group and count genes by status
summary_table <- gene_overlap_table %>%
  group_by(Status) %>%
  summarize(UniqueGenes = n(), .groups = "drop")

# Display the summary table
print(summary_table)

```
