---
title: "SS3x"
author: "Domi"
date: "2025-01-16"
output: html_document
editor_options: 
  chunk_output_type: console
---
Introduction
This script is dedicated to visualizing preprocessed data from **1DP_01_mg_cells_raw_data_to_seurat_merged_integrated_UMAP.Rmd**. The processed data are represented using **UMAP plots, dot plots, and stacked plots** to illustrate clustering, marker expression, and population distributions. These visualizations provide insights into cell-type composition and gene expression patterns across different isolation protocols.


```{r Load packages and lists,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
suppressMessages(library("Matrix"))
suppressMessages(library("RColorBrewer"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("sctransform"))
suppressMessages(library("scales"))
suppressMessages(library("dplyr"))
suppressMessages(library("SeuratWrappers"))
suppressMessages(library("reshape2"))
suppressMessages(library("ComplexHeatmap"))
suppressMessages(library("gridExtra"))
suppressMessages(library("magrittr"))
suppressMessages(library("patchwork"))
suppressMessages(library("ggpubr"))
plots <- list()
col_list <- list()
orders <- list()
source("0F_general_functions.R", local = F, echo = F, print.eval = F)
source("0F_colours_orders.R", local = F, echo = F, print.eval = F)
#Read table with gene symbols end IDs
gene_symbol <- read.table("data/Smartseq3xpress.gene_names.txt", header = TRUE)
```

#UMAP visualisation
```{r seurat LOAD,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#Load seurat
seurat_mg <- readRDS("ws/1DP_01_seurat_merged_integrated_UMAP_mg_cells.rds")
```

```{r UMAP_all_microglia,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
plots[["UMAP_all_mg"]] <- DimPlot(
  seurat_mg,
  reduction = "umap.cca_rm",
  group.by = "Cell_types_mg_abbr",
  pt.size = 0.1,  
  cols = col_list[["UMAP_mg"]]
) + 
  ggtitle("All conditions") + 
 ggtitle("All conditions") + 
  theme_umap +
  remove_grid +
  remove_axes+
  guides(colour = guide_legend(override.aes = list(size = 1)))

# Display the plot
plots[["UMAP_all_mg"]]

```

```{r UMAP_all_cells_condition_LiveCells,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_filtered_LiveCells <- subset(seurat_mg, LiveCells != "LiveNuclei + FixedNuclei")

plots[["UMAP_mg_LiveCells"]] <- DimPlot(
  seurat_filtered_LiveCells,
  reduction = "umap.cca_rm",
  group.by = "LiveCells",
  pt.size = 0.1,
  cols = col_list[["UMAPLiveCells"]]
) + 
  ggtitle("LiveCells") + 
  theme_umap +
  remove_grid +
  remove_axes+
  guides(colour = guide_legend(override.aes = list(size = 1)))


plots[["UMAP_mg_LiveCells"]]
#rm(seurat_filtered_LiveCells)
```

```{r UMAP_all_cells_condition_LiveNuclei,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_filtered_LiveNuclei <- subset(seurat_mg, LiveNuclei != "LiveCells + FixedNuclei")


plots[["UMAP_mg_LiveNuclei"]] <- DimPlot(
  seurat_filtered_LiveNuclei,
  reduction = "umap.cca_rm",
  group.by = "LiveNuclei",
  pt.size = 0.1,
  cols = col_list[["UMAPLiveNuclei"]]
) + 
  ggtitle("LiveNuclei") + 
 theme_umap +
  remove_grid +
  remove_axes+
  guides(colour = guide_legend(override.aes = list(size = 1)))


plots[["UMAP_mg_LiveNuclei"]]
#rm(seurat_filtered_LiveNuclei)

```

```{r UMAP_all_cells_condition_FixNuclei,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
seurat_filtered_FixedNuclei <- subset(seurat_mg, FixedNuclei != "LiveCells + LiveNuclei")


plots[["UMAP_mg_FixedNuclei"]] <- DimPlot(
  seurat_filtered_FixedNuclei,
  reduction = "umap.cca_rm",
  group.by = "FixedNuclei",
  pt.size = 0.1,
  cols = col_list[["UMAPFixedNuclei"]]
) + 
  ggtitle("FixedNuclei") + 
  theme_umap +
  remove_grid +
  remove_axes+
  guides(colour = guide_legend(override.aes = list(size = 1)))


plots[["UMAP_mg_FixedNuclei"]]
#rm(seurat_filtered_FixedNuclei)
```

```{r Table with cells, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
#info about cells per condition
table(seurat_mg@meta.data$Cond_preparation)
table(seurat_mg@meta.data$Cell_types_mg_abbr)
```

# Fig. 02 A UMAP visualisation in combination of 4 UMAPs
```{r UMAP saving, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
layout_1stline <- "
ABCD
"
plots[["1st_line02"]] <- plots[["UMAP_all_mg"]] + plots[["UMAP_mg_LiveCells"]] + plots[["UMAP_mg_LiveNuclei"]] + plots[["UMAP_mg_FixedNuclei"]] +  
   plot_layout(design =  layout_1stline,
              widths = c(1,1,1,1)) +
  plot_annotation(
    tag_levels = list(c('A', ''),  
                      theme = theme(
                        plot.tag = element_text(size = 7, face = "plain")  # Customize tag size and appearance
                      )))

plots[["1st_line"]]

ggsave(
  filename = "results/2V_Fig02_1stlineUMAP.svg",              
  plot = plots[["1st_line02"]],          
  device = "svg",                          
  width = 16.9,                            
  height = 5,                              
  units = "cm"                            
)
```


#Dotplot visualisation of genes
```{r List of marker_genes_all,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
markers_mg <- c("P2ry12", "Tmem119", "Hexb",  #Homeostatic
                "C1qa", "C1qc", "C1qb", #Reactive
                "Trem2","Lyz2","Apoe", "Cst7", "Lpl", "Clec7a",    #ARM
                "Ccl6", "Ccl3", "Ccl4",
                "Ifit3",  "Irf7", "Ifitm3", #IRM
                "Birc5",  "Tuba1b", "Stmn1"  #Dividing
                
)
```

```{r RNA normalization,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
DefaultAssay(seurat_mg) <- "RNA" 
seurat_mg <- JoinLayers(seurat_mg, assay = "RNA")
all.genes.RNA <- rownames(seurat_mg)
seurat_mg <- NormalizeData(seurat_mg, normalization.method = "LogNormalize", scale.factor = 10000)
seurat_mg <- ScaleData(seurat_mg, features = all.genes.RNA)
```

```{r Dot plot with feature test,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
DefaultAssay(seurat_mg) <- "RNA"
Idents(seurat_mg) <- "Cell_types_mg_abbr"


plots[["dot_plot_mg"]] <- DotPlot(
  seurat_mg, 
  features = Symbol_to_ENSEMBL(markers_mg), 
 dot.min = 0.20 
) + 
  scale_x_discrete(labels = ENSEMBL_to_Symbol(Symbol_to_ENSEMBL(markers_mg))) + 
  theme_mk + 
  remove_grid +
  theme(
    axis.text.x = element_text(angle = 290, hjust = 0, vjust = 0.5, size = 6, face = "italic"),
    axis.title.x = element_blank(),                                
    axis.title.y = element_blank(),                               
    legend.position = "bottom",                                    
    plot.margin = margin(0,0,0,0),                                  
  legend.margin = margin(b = -5, l = -3),
  legend.box.spacing = unit(0.1, "cm"),
  legend.box.just = "left"
    ) +
  scale_color_gradientn(
    name = "Scaled average \ngene expression",  
    colors = c("#E8ECEF","#9ABDDC", "#3A0956")             
  ) +
  scale_size_continuous(
    name = "Pergentage \nof cells (%)",
    range = c(0.5, 2)              # Set the range for dot sizes; adjust as needed
  ) +
  guides(
    color = guide_colorbar(title.position = "left"), # Adjust legend color bar position if needed
    size = guide_legend(title.position = "left")     # Adjust size legend position if needed
  )
  

plots[["dot_plot_mg"]]
```

# Fig 02 B
```{r Fig02_2ndline,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
layout_2ndline <- "
A
"

plots[["2nd_line_dot"]] <- plots[["dot_plot_mg"]] +
  plot_layout(design =  layout_2ndline) +
  plot_annotation(
    tag_levels = list(c('B'),  
                      theme = theme(
                        plot.tag = element_text(size = 7, face = "plain") 
                      )))
                      
plots[["2nd_line_dot"]]

ggsave(
  filename = "results/2V_Fig02_2ndline_dot.svg",             
  plot = plots[["2nd_line_dot"]],          
  device = "svg",                         
  width = 9.35,                             
  height = 5,                              
  units = "cm"                            
)

```


#Stacked bar_plots with relative proportion of cells in clusters
```{r Data frame with percentages for clusters per condition,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
metadata_df <- read.csv("ws/1DP_01_mg_seurat_metadata.csv")
# Calculate total counts per condition
total_counts_per_condition <- metadata_df %>%
  group_by(Cond_publication) %>%
  tally(name = "total_cells")

# Calculate counts per cluster per condition
cluster_counts <- metadata_df %>%
  group_by(Cond_publication, Cell_types_mg_abbr) %>%
  tally(name = "cluster_count")

# Join the total counts with cluster counts
perc_clust_per_condition <- cluster_counts %>%
  left_join(total_counts_per_condition, by = "Cond_publication") %>%
  mutate(percent = (cluster_count / total_cells) * 100) %>%
  select(Cond_publication, Cell_types_mg_abbr, percent)

# View the resulting data frame
print(perc_clust_per_condition)

rm(total_counts_per_condition, cluster_counts)

#change order of cell types - inverted
perc_clust_per_condition$Cell_types_mg_abbr <- factor(perc_clust_per_condition$Cell_types_mg_abbr, levels = orders[["Cell_types_mg_abbr_invert"]])
perc_clust_per_condition$Cell_types_mg_abbr %>% table()

#group Sham together and MCAO 
perc_clust_per_condition$Cond_publication <- factor(perc_clust_per_condition$Cond_publication, levels = orders[["Cond_publication"]])



# Group microglia cluster into one - better for visualisation 
perc_clust_per_condition %<>% mutate("Cond_SHAMvsMCAO" = factor(case_when(
   Cond_publication == "LiveCells_Sham" ~ "Sham",
   Cond_publication == "LiveNuclei_Sham"~ "Sham",
   Cond_publication == "FixedNuclei_Sham" ~ "Sham",
   Cond_publication ==  "LiveCells_MCAO"~ "MCAO",
   Cond_publication == "LiveNuclei_MCAO" ~ "MCAO",
   Cond_publication == "FixedNuclei_MCAO" ~ "MCAO"
)))

#group SHAM or MCAO
perc_clust_per_condition$Cond_SHAMvsMCAO <- factor(perc_clust_per_condition$Cond_SHAMvsMCAO, levels = orders[["Cond_SHAMvsMCAO"]])


perc_clust_per_condition %<>% mutate("Cond_appr" = factor(case_when(
   Cond_publication == "LiveCells_Sham" ~ "LiveCells",
  Cond_publication == "LiveNuclei_Sham"~ "LiveNuclei",
  Cond_publication == "FixedNuclei_Sham" ~ "FixedNuclei",
  Cond_publication ==  "LiveCells_MCAO"~ "LiveCells",
  Cond_publication == "LiveNuclei_MCAO" ~ "LiveNuclei",
  Cond_publication == "FixedNuclei_MCAO" ~ "FixedNuclei"
)))
```

```{r Stacked bar plot - condition vs percentage of cells,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
plots[["stacked_bar_mg"]] <- ggplot(perc_clust_per_condition, aes(x = Cond_appr, y = percent, fill = Cell_types_mg_abbr)) +
  geom_bar(position = "stack", stat = "identity", color = "#2d1800", size = 0.3 ) + 
  scale_fill_manual(values = col_list[["UMAP_mg"]]) +  
  labs(x = "Conditions", y = "Relative proportion of cells (%)", fill = "Cell \ntypes") +
 theme_mk +
  scale_y_continuous(
    expand = c(0, 0),               # Remove extra padding
    breaks = seq(0, 100, by = 20))+  # Set y-axis breaks at 20-unit intervals
  facet_grid(. ~ Cond_SHAMvsMCAO)+
  remove_grid +
   theme(
    strip.text = element_text(size = 6),        # Customize strip text size
    strip.background = element_blank(),         # Remove the facet strip background
    legend.position = "right",
    legend.margin = margin(t = -7.5),
    plot.margin = margin(0,0,0,0),
    axis.text.x = element_text(angle = 340, vjust = +0.1, hjust = +0.2),
    axis.title.x = element_blank(), 
    legend.text = element_text(margin = margin(t = 0, r = 1, b = 1, l = 1)),
    legend.title = element_text(margin = margin(t = 0, r = 1, b = 6, l = 1))
    )  


plots[["stacked_bar_mg"]]
```

# Fig 02 C
```{r Fig02_2nd line StackedBar Plot,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
ggsave(
  filename = "results/2V_Fig02_2ndline_stack.svg",             
  plot = plots[["stacked_bar_mg"]],        
  device = "svg",                          
  width = 7,                             
  height = 4.5,                             
  units = "cm"                            
)
```


#ARM, IRM expression from Frigerio2019
```{r}
ARM_up <-  Symbol_to_ENSEMBL(c("Apoe", "Cst7", "Lpl", "Clec7a", "Lyz2", "Ccl3", "Lgals3bp", "Cd63", "H2-D1", "Ctsb"))
seurat_mg <- AddModuleScore(seurat_mg, features = list(ARM_up), assay = "RNA", name = "ARM_up_")

ARM_down <-  Symbol_to_ENSEMBL(c("Selplg", "Cd164", "P2ry12", "Crybb1", "Tmem119", "Stab1", "Ivns1abp", "Ecscr", "P2ry13", "Slc2a5"))
seurat_mg <- AddModuleScore(seurat_mg, features = list(ARM_down), assay = "RNA", name = "ARM_down_")

IRM_up <-  Symbol_to_ENSEMBL(c("Ifit3", "Ifitm3", "Rtp4", "Irf7", "Oasl2", "Ifi204",  "Usp18", "Ifi209", "Ifi27l2a", "Oas1a"))
seurat_mg <- AddModuleScore(seurat_mg, features = list(IRM_up), assay = "RNA", name = "IRM_up_")

CPM_up <-  Symbol_to_ENSEMBL(c("Mcm5", "Mcm2", "Mcm6", "Lig1", "Stmn1", "Tipin", "Hmgb2", "Dnmt1", "Cbx5", "Mcm7"))
seurat_mg <- AddModuleScore(seurat_mg, features = list(CPM_up), assay = "RNA", name = "CPM_up_")
```


```{r Feature plots from Frigerio,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Define the features and their corresponding titles
features <- c("ARM_up_1", "ARM_down_1", "IRM_up_1", "CPM_up_1")
titles <- c("ARM \n(Sala Frigerio et al, 2019)", "HM \n(Sala Frigerio et al, 2019)", "IRM \n(Sala Frigerio et al, 2019)", "CPM \n(Sala Frigerio et al, 2019)")  # Custom titles

# Define custom breaks for each feature
custom_breaks <- list(
  ARM_up_1 = c(0.5, 1.2),    # Example: min, mid, max values for "ARM_up_1"
  ARM_down_1 = c(0.5,1.0),  # Example: min, mid, max values for "ARM_down_1"
  IRM_up_1 = c(0.25, 0.75),   # Example: min, mid, max values for "IRM_up_1"
  CPM_up_1 = c(0.08, 0.18)   # Example: min, mid, max values for "CPM_up_1"
)


# Loop over each feature to create the plots
for (i in seq_along(features)) {
  feature <- features[i]  # Get the current feature
  plot_title <- titles[i]  # Get the corresponding custom title
  
  # Get custom breaks and labels for the current feature
  breaks <- custom_breaks[[feature]]
  labels <- custom_breaks[[feature]]

  # Create the plot
  plots[[paste0(feature, "_feature")]] <- FeaturePlot(
    seurat_mg,
    features = feature,
    min.cutoff = "q10",
    max.cutoff = "q90",
    reduction = "umap.cca_rm", 
    pt.size = 0.1
  ) +
    ggtitle(plot_title) + 
    theme_umap +
    remove_axes+
    remove_grid +
    theme(
      legend.position = "bottom",             
      legend.title = element_text(size = 6),  
      legend.text = element_text(size = 6),   
     legend.margin = margin(t = -10),
      plot.margin = margin(0, 0, 4, 0),
      legend.key.size = unit(2.5, "mm")
    ) +
    scale_color_gradientn(
      name = "Average \nExpression",          
      colors = c("#c6daeb", "#3A0956"),
      breaks = breaks,                        # Use manually set breaks
      labels = labels                         # Use manually set labels
    )
}
plots[["feature"]] <- plots[["ARM_down_1_feature"]] + plots[["ARM_up_1_feature"]] +  plots[["IRM_up_1_feature"]] + plots[["CPM_up_1_feature"]] 
plots[["feature"]]
```

# Fig 02 D
```{r Fig02 Feature Plot,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
layout_3rdline <- "
ABCD
"
plots[["3rd_line"]] <- plots[["ARM_down_1_feature"]] + plots[["ARM_up_1_feature"]] +  plots[["IRM_up_1_feature"]] + plots[["CPM_up_1_feature"]] +
   plot_layout(design = layout_3rdline) +
   plot_annotation(
    tag_levels = list(c('D', ''),  
                      theme = theme(
                        plot.tag = element_text(size = 7, face = "plain")  
                      )))


plots[["3rd_line"]]

ggsave(
  filename = "results/2V_Fig02_3rdline_feature.svg",           
  plot = plots[["3rd_line"]],          
  device = "svg",                          
  width = 16.9,                             
  height = 5.5,                              
  units = "cm"                            
)
```

# Violin plots showing the expression of marker genes from Frigerio et al. (2019) in our dataset
```{r Violin_plots,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Define the features for which you want to create violin plots
features <- c("ARM_up_1", "ARM_down_1", "IRM_up_1", "CPM_up_1")
titles <- c(
  "ARM \n(Sala Frigerio et al, 2019)", 
  "HM \n(Sala Frigerio et al, 2019)", 
  "IRM \n(Sala Frigerio et al, 2019)", 
  "CPM \n(Sala Frigerio et al, 2019)"
)

# Define your comparisons
comparisons <- list(
  "ARM_up_1" = list(c("HM1", "ARM"), c("HM2", "ARM") ),
  "ARM_down_1" = list(c("HM1", "ARM"), c("HM2", "ARM")),
  "IRM_up_1" = list(c("HM1", "IRM"), c("HM2", "IRM")),
  "CPM_up_1" = list(c("HM1", "CPM"), c("HM2", "CPM"))
)
# Loop over each feature to create the plots
for (i in seq_along(features)) {
  feature <- features[i]  # Get the current feature
  plot_title <- titles[i]
  feature_comparisons <- comparisons[[feature]]  # Get the comparisons for the current feature
  
  # Calculate the maximum y-value in the data for scaling
  max_y <- max(seurat_mg@meta.data[[feature]], na.rm = TRUE)
  
  # Define custom label.y values for each comparison
  label_y_positions <- c(
    0.8 * max_y,  # For the first comparison
    0.9* max_y   # For the second comparison
  )
  
  # Create the plot
  plots[[paste0(feature, "_violin")]] <- VlnPlot(
    seurat_mg,
    features = feature,
    group.by = "Cell_types_mg_abbr",
    pt.size = 0.0,  # Adjust this value if needed
    cols = col_list[["UMAP_mg"]]
  ) +
    ggtitle(plot_title) + 
    theme_mk_title +
    remove_grid +
    theme(
      axis.text.x = element_text(angle = 290, hjust = 0, vjust = 0.5), 
      legend.position = "bottom",             
      legend.title = element_text(size = 6),  
      legend.text = element_text(size = 6),   
      legend.margin = margin(t = -10),
      plot.margin = margin(0, 0, 2, 0),
      axis.title.x = element_blank()          
    ) +
    NoLegend() +
    # Overlay box plot with white fill
    geom_boxplot(width = 0.1, outlier.shape = NA, color = "black", fill = "white") +
    labs(y = if (i == 2) "Average Expression" else NULL) +
    # Add statistical comparison with adjusted spacing
    stat_compare_means(
      comparisons = feature_comparisons,                     # Use the comparisons specific to this feature
      method = "wilcox.test",                                # Wilcoxon test
      label = "p.signif",                                    # Display significance
      method.args = list(paired = FALSE),                   # Specify unpaired test
      hide.ns = FALSE,                                      # Display non-significant results
      label.x.npc = "left",                                # Position labels horizontally
      label.y = label_y_positions,                          # Use custom label.y positions
      vjust = 0.5                                        # Adjust the vertical position of asterisks
    )
}

# Combine the plots (example)
plots[["violin"]] <- plots[["ARM_down_1_violin"]] + plots[["ARM_up_1_violin"]] +
  plots[["IRM_up_1_violin"]] + plots[["CPM_up_1_violin"]]
plots[["violin"]]
```

# Fig 02 E
```{r Fig02_4th line Violin Plot,  eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
layout_4thline <- "
ABCD
"
plots[["4th_line"]] <- plots[["ARM_down_1_violin"]] + plots[["ARM_up_1_violin"]] +  plots[["IRM_up_1_violin"]] + plots[["CPM_up_1_violin"]] +
     plot_layout(design = layout_3rdline) +
   plot_annotation(
    tag_levels = list(c('E', ''),  
                      theme = theme(
                        plot.tag = element_text(size = 7, face = "plain")  
                      )))


plots[["4th_line"]]

ggsave(
  filename = "results/2V_Fig02_4thline_violin.svg",           
  plot = plots[["4th_line"]],        
  device = "svg",                          
  width = 16.9,                              
  height = 5.5,                              
  units = "cm"                            
)
```